---
title: "Analysis_0.3_Taxonomic_groups"
author: "Charles T. Bond"
date: "`r Sys.Date()`"
output: html_document
---


# Microbial Community Analysis: Talladega Forest Non-perennial Stream Network

## Analysis_0.3: Microbial taxa across the stream network

### Introduction
Following the bioinformatics processing and Analysis 0.1 and 0.2, we will now look into the taxonomic groups of prokaryotes and fungi occuring in the stream. See the next analysis (0.4) for generalized linear latent variable models for top genera, species, or PICRUSt2-based functional trait assignment.

## Setup
```{r, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/chunk/Talladega_synoptic/Analysis")
knitr::opts_knit$set(root.dir = "/Users/chunk/Talladega_synoptic/Analysis")
options(knitr.duplicate.label = "allow") ### so I don't have to give chunks unique names
```

```{r}
#library(dada2)
library(phyloseq)
library(vegan)
library(ggplot2)
library(tidyr)
library(dplyr)
library(pheatmap)
library(GUniFrac)
library(pals)
library(RColorBrewer)
library(ragg)
library(ggpubr)

#Set theme
theme_set(theme_bw() + theme(
              plot.title = element_text(size=20, color="black"),
              axis.text.x = element_text(size=15, color="black"),
              axis.text.y = element_text(size=15, color="black"),
              axis.title.x = element_text(size=15),
              axis.title.y = element_text(size=15),
              legend.text = element_text(size=12),
              legend.title = element_text(size=15),
            #  legend.position = "bottom",
            #  legend.key=element_blank(),
            #  legend.key.size = unit(0.5, "cm"),
            #  legend.spacing.x = unit(0.1, "cm"),
            #  legend.spacing.y = unit(0.1, "cm"),
              panel.background = element_blank(), 
              #panel.border = element_rect(colour = "black", fill=NA, size=1),
              plot.background = element_blank()))
```

## Load phyloseq objects
```{r}
#prokaryotes
load("/Users/chunk/Talladega_synoptic/Analysis/Rdata/pseqtest16S.rdata")
pseqtest16S

#fungi
load("/Users/chunk/Talladega_synoptic/Analysis/Rdata/pseqtestITS.rdata")
pseqtestITS

## Cutting the single beaver pond site, its an outlier, the only permanent lentic site
pseqtest16S <- subset_samples(pseqtest16S, siteId != "TLMBP")
pseqtestITS <- subset_samples(pseqtestITS, siteId != "TLMBP")

# extract sampe data
samdf_16S <- data.frame(sample_data(pseqtest16S))
samdf_ITS <- data.frame(sample_data(pseqtestITS))
```

## 0.3.1 
### Relative Abundances of Major Prokaryotic Taxa
First we will process and visualize the relative abundances of major taxonomic groups, and test a few hypotheses about our major taxa. 

#### 0.3.1.1
##### Prokaryotes: Bacteria vs. Archaea
We want to compare the relative abundance of Bacteria and Archaea in order to test the hypothesis that 
1) relative abundance of Archaea is higher in sediments than in other substrates, and 
2) that the relative abundance of Archaea is positively correlated with water permanence

```{r}
kingcount16S <- otu_table(tax_glom(pseqtest16S, taxrank="Kingdom"), taxa_are_rows = FALSE)
kingcount16S <- t(kingcount16S)
#make vector of phyla names to set as row names
kingcount16S_vec <- as.vector(tax_table(tax_glom(pseqtest16S, taxrank="Kingdom"))[,1]) 
rownames(kingcount16S) <- as.vector(kingcount16S_vec)
## convert to percent
kingprop16S <- apply(kingcount16S, 2, function(x) x/sum(x)*100)

### merge with metadata
kingmerge <- merge(samdf_16S, t(kingprop16S),
                  by="row.names", all=TRUE)
rownames(kingmerge) <- kingmerge[,1]
kingmerge <- kingmerge[,-1]

### Overall Kingdom percentages
mean(kingmerge$Bacteria)
mean(kingmerge$Archaea)



library(rstatix)
stat.test <- kingmerge %>%
wilcox_test(Archaea~substrate) %>%
  adjust_pvalue(method = "bonferroni") %>%
  add_significance()
stat.test
effsize <- kingmerge %>%
  wilcox_effsize(Archaea~substrate)
effsize

cor_summary <- kingmerge[!is.na(kingmerge$percentwet_11month),] %>%
  group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Archaea, percentwet_11month, method = "spearman"),
    p_value = cor.test(Archaea, percentwet_11month, method = "spearman")$p.value
  )
cor_summary

cor_summary <- kingmerge[!is.na(kingmerge$CH4_avg),] %>%
  group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Archaea, CH4_avg, method = "spearman"),
    p_value = cor.test(Archaea, CH4_avg, method = "spearman")$p.value
  )
cor_summary
```

#### Prokaryote phyla and classes

```{r}
#make taxonomy object by phylum
phyla16S_counts_tab <- otu_table(tax_glom(pseqtest16S, taxrank="Phylum"), taxa_are_rows = FALSE)
phyla16S_counts_tab <- t(phyla16S_counts_tab)
#make vector of phyla names to set as row names
phyla16S_tax_vec <- as.vector(tax_table(tax_glom(pseqtest16S, taxrank="Phylum"))[,2]) 
rownames(phyla16S_counts_tab) <- as.vector(phyla16S_tax_vec)
phyla16S_counts_tab
write.csv(phyla16S_counts_tab, 'phyla16S_counts_05.21.25.csv')
          
```

```{r}
asv_counts <- pseqtest16S@otu_table
#determine the number of unclassified seqs at the phylum level
unclassified_tax_counts <- colSums(t(asv_counts)) - colSums(phyla16S_counts_tab)
#Add a row of "unclassified" to the phylum count table
phyla16S_and_unidentified_counts_tab <- rbind(phyla16S_counts_tab, "Unclassified_Phylum"=unclassified_tax_counts)

```

### split Pseudomonadota (Proteobacteria) into classes
Apparently, in the time since I updated SILVA, Proteobacteria has been changed to 'Pseudomonadota', so I've had to update this code a bit. 
```{r}
#remove Proteobacteria for the purpose of separating it into classes later
temp_major16S_taxa_counts_tab <- 
  phyla16S_and_unidentified_counts_tab[!row.names(phyla16S_and_unidentified_counts_tab)
                                    %in% "Pseudomonadota",]
#make count table broken down by class
class16S_counts_tab <- otu_table(tax_glom(t(pseqtest16S), taxrank="Class"))
#make table containing phylum and class data
class16S_tax_phy_tab <- tax_table(tax_glom(t(pseqtest16S), taxrank="Class"))
phy_tmp_vec <- class16S_tax_phy_tab[,2]
class16S_tmp_vec <- class16S_tax_phy_tab[,3]
rows_tmp <- row.names(class16S_tax_phy_tab)
class16S_tax_tab <- data.frame("Phylum"=phy_tmp_vec, "Class"=class16S_tmp_vec, row.names = rows_tmp)
#make vector of just the Proteobacteria classes
Pseudomonad_classes_vec <- as.vector(class16S_tax_tab[class16S_tax_tab$Phylum=="Pseudomonadota", "Class"])
row.names(class16S_counts_tab) <- as.vector(class16S_tax_tab$Class)
#make table of Proteobacteria classes
Pseudomonad_class_counts_tab <- class16S_counts_tab[row.names(class16S_counts_tab) %in% Pseudomonad_classes_vec, ]
#make table ofProteobacteria not identified to the class level
Pseudomonad_no_class_annotated_counts <- 
  phyla16S_and_unidentified_counts_tab[row.names(phyla16S_and_unidentified_counts_tab) %in% "Pseudomonadota",]-
  colSums(Pseudomonad_class_counts_tab)
#Now combine the tables
major16S_taxa_counts_tab <- rbind(temp_major16S_taxa_counts_tab, Pseudomonad_class_counts_tab,
                               "Unclassified_Pseudomonadota"=Pseudomonad_no_class_annotated_counts)
head(major16S_taxa_counts_tab)

```

```{r}
write.csv(major16S_taxa_counts_tab, "major16Staxa_count_rar_16S_05.21.25.csv")
#Check that all sequences are accounted for
identical(colSums(major16S_taxa_counts_tab), rowSums(asv_counts))
## [1] TRUE
#Convert totals to relative abundance
major16S_taxa_proportions_tab <- apply(major16S_taxa_counts_tab, 2, function(x) x/sum(x)*100)
#colSums(major16S_taxa_proportions_tab)
write.csv(major16S_taxa_proportions_tab, "major16Staxa_relative abundance_rar_16S_05.21.25.csv")

```

by family
```{r}
#make taxonomy object by genus
family16S_counts_tab <- otu_table(tax_glom(pseqtest16S, taxrank="Family"), taxa_are_rows = FALSE)
family16S_counts_tab <- t(family16S_counts_tab)
#make vector of genus names to set as row names
family16S_tax_vec <- as.vector(tax_table(tax_glom(pseqtest16S, taxrank="Family"))[,5]) 
rownames(family16S_counts_tab) <- as.vector(family16S_tax_vec)

#determine the number of unclassified seqs at the family level
unclassified_family16S_counts <- colSums(t(asv_counts)) - colSums(family16S_counts_tab)
unclassified_family16S_counts
#Add a row of "unclassified" to the family count table
family16S_and_unidentified_counts_tab <- rbind(family16S_counts_tab, 
                                            "Unclassified_family"=unclassified_family16S_counts)
write.csv(family16S_and_unidentified_counts_tab, "family16S raw abundance_rar_16S_05.21.25.csv")
#Check that all seqs are accounted for.

identical(colSums(family16S_and_unidentified_counts_tab), rowSums(asv_counts))

#Convert totals to relative abundance
family16S_proportions_tab <- apply(family16S_and_unidentified_counts_tab, 2, function(x) x/sum(x)*100)
write.csv(family16S_proportions_tab, "family relative abundance_rar_16S_05.21.25.csv")

```


```{r}
#Merge metadata, phylum, and family abundances
#phy_and_fam <- merge(t(major16S_taxa_counts_tab), t(family_and_unidentified_counts_tab),
#                  by="row.names", all=TRUE)
phy_and_fam <- merge(t(major16S_taxa_proportions_tab), t(family16S_proportions_tab),
                  by="row.names", all=TRUE)
phyfam <- phy_and_fam[,-1]
rownames(phyfam) <- phy_and_fam[,1]
tax_merge <- merge(samdf_16S, phyfam,
                  by="row.names", all=TRUE)
taxmerge16S <- tax_merge[,-1]
rownames(taxmerge16S) <- tax_merge[,1]
write.csv(tax_merge, "TALSYN_taxmerge16S_05.21.25.csv")

```


###Phylum-Class BARPLOT
```{r}
#major16S taxa, mean percent abundances across all samples:
rowMeans(major16S_taxa_proportions_tab)

#Select taxa that make up >90% of the total sequences

phyla16Strim <- data.frame(major16S_taxa_proportions_tab[c("Alphaproteobacteria","Gammaproteobacteria","Cyanobacteriota","Actinomycetota","Myxococcota","Verrucomicrobiota","Bacteroidota","Acidobacteriota","Bacillota","Planctomycetota","Thermodesulfobacteriota","Chloroflexota","Thermoproteota"), ])
#write.csv(phylatrim_t,"/Users/chunk/Documents/DADA2/DADA2_package_test/Kz_Syn_ITS/run_2/outputs/phylatrim.csv")
#write.csv(major16S_taxa_proportions_tab,"/Users/chunk/Documents/DADA2/DADA2_package_test/Kz_Syn_ITS/major16Staxa.csv")
#major16Staxa <- read.csv("~/Documents/DADA2/DADA2_package_test/Kz_Syn_ITS/major16Staxa.csv", row.names=1)
#View(major16Staxa)
#phylatrim <- read.csv("~/Documents/DADA2/DADA2_package_test/Kz_Syn_ITS/phylatrimt.csv", row.names=1)
#View(phylatrim)

#Create an "other" category for the phyla not retained
filtered_proportions <- colSums(major16S_taxa_proportions_tab) - 
  colSums(phyla16Strim)
phyla16Strim <- rbind(phyla16Strim, "Other"=filtered_proportions)
phyla16Strim

#Add taxa names as a column
phyla16Strim$Major_Taxa <- row.names(phyla16Strim)
#transform into long format
phyla16Slong <- gather(phyla16Strim, Sample, Proportion, -Major_Taxa)
phyla16Slong$Sample <- gsub("X","",phyla16Slong$Sample)
#add metadata
#metadatap <- read.csv("~/Documents/DADA2/DADA2_package_test/Kz_Syn_ITS/metadata_punc.csv", row.names=1, stringsAsFactors=TRUE)
#View(metadata_punc)
#names(metadatap)
```

```{r}
## [1] "substrate" "WetDry"
phyla16Smet<-data.frame("Sample"=row.names(samdf_16S),
                     "Substrate"=samdf_16S$substrate,
                     "Classiification"=samdf_16S$class_state,
                     stringsAsFactors=T)
#merge metadata with major taxa data
phyla16Slong <- merge(phyla16Slong, phyla16Smet)
#Summarize by depth and hydration
phyla16S_summary <- 
  phyla16Slong %>% # the names of the new data frame and the data frame to be summarised
  group_by(Substrate, Classiification, Major_Taxa) %>%   # the grouping variable
  summarise(mean_prop = mean(Proportion))  # calculates the mean of each group

phyla16S_summary

phyla16S_summary$Substrate <- factor(phyla16S_summary$Substrate,
                                 levels=c("L","B","S")) #reorder variables

#phyla16S_summary$Wet.Dry <- factor(phyla16S_summary$Wet.Dry, levels= c("wet","dry")) #reorder variables
phyla16S_summary$Major_Taxa <- factor(phyla16S_summary$Major_Taxa, levels= rev(c("Alphaproteobacteria","Gammaproteobacteria","Cyanobacteriota","Actinomycetota","Myxococcota","Verrucomicrobiota","Bacteroidota","Acidobacteriota","Bacillota","Planctomycetota","Thermodesulfobacteriota","Chloroflexota", "Thermoproteota","Other")))

#color palette
sublab<- c("Leaf litter","Epilithic biofilms","Benthic sediment")
names(sublab)<- c("L","B","S")

pal12 <- rev(cols25(n=14)) #set n= to the number of taxa you are plotting
#make stacked bar plot
phylum16S_bar <- ggplot(phyla16S_summary, aes(x = Classiification, y = mean_prop, fill = Major_Taxa))+
  geom_bar(stat = "identity", col=I("black")) +
  scale_fill_manual(values=pal12)+
  guides(fill=guide_legend(ncol=1))+
  facet_wrap(~Substrate, labeller = labeller(Substrate=sublab), nrow=1, scales="free_x") +
  labs(x=NULL,y="Relative abundance (%)",
       fill="Major taxa")+
  #scale_facet_discrete(labels=c('B'='Epilithic biofilms', 'L'='Leaf litter', 'S'='Benthic sediment', 'W'='Surface water'))+
  theme(axis.text.x = element_text(angle=60, hjust=1), legend.position = "right")+
  theme(text=element_text(size=18), #change font size of all text
        axis.text=element_text(size=16), #change font size of axis text
        axis.title=element_text(size=16), #change font size of axis titles
        plot.title=element_text(size=16), #change font size of plot title
        legend.text=element_text(size=16), #change font size of legend text
        strip.text.x = element_text(size = 16),
        legend.title=element_text(size=16)) #change font size of legend title  
phylum16S_bar
 
#c(180,100) *
#  0.0394 * # convert mm to inch
#  600 # convert to pixels
## [1] 4255.2 2364.0
plotout <- "phylum16S_barplot__052125.tiff"
agg_tiff(filename=plotout, width=4255, height=2364, units="px",
         pointsize=10, res=600, compression="lzw", scaling=0.5)
phylum16S_bar
invisible(dev.off())
```
##################################################

##################################################


#### 0.3.1.2
### Fungal Taxonomy
#### Fungal phyla and classes
```{r}
#make taxonomy object by phylum
phylaITS_counts_tab <- otu_table(tax_glom(pseqtestITS, taxrank="Phylum"), taxa_are_rows = FALSE)
phylaITS_counts_tab <- t(phylaITS_counts_tab)
#make vector of phyla names to set as row names
phylaITS_tax_vec <- as.vector(tax_table(tax_glom(pseqtestITS, taxrank="Phylum"))[,2]) 
rownames(phylaITS_counts_tab) <- as.vector(phylaITS_tax_vec)
phylaITS_counts_tab
write.csv(phylaITS_counts_tab, 'phylaITS_counts_05.21.25.csv')
          
```

```{r}
asv_counts <- pseqtestITS@otu_table
#determine the number of unclassified seqs at the phylum level
unclassified_tax_counts <- colSums(t(asv_counts)) - colSums(phylaITS_counts_tab)
#Add a row of "unclassified" to the phylum count table
phylaITS_and_unidentified_counts_tab <- rbind(phylaITS_counts_tab, "Unclassified_Phylum"=unclassified_tax_counts)

```

### split Ascomycota

```{r}
#remove Proteobacteria for the purpose of separating it into classes later
temp_majorITS_taxa_counts_tab <- 
  phylaITS_and_unidentified_counts_tab[!row.names(phylaITS_and_unidentified_counts_tab)
                                    %in% "Ascomycota",]
#make count table broken down by class
classITS_counts_tab <- otu_table(tax_glom(t(pseqtestITS), taxrank="Class"))
#make table containing phylum and class data
classITS_tax_phy_tab <- tax_table(tax_glom(t(pseqtestITS), taxrank="Class"))
phy_tmp_vec <- classITS_tax_phy_tab[,2]
classITS_tmp_vec <- classITS_tax_phy_tab[,3]
rows_tmp <- row.names(classITS_tax_phy_tab)
classITS_tax_tab <- data.frame("Phylum"=phy_tmp_vec, "Class"=classITS_tmp_vec, row.names = rows_tmp)
#make vector of just the Proteobacteria classes
Ascomycete_classes_vec <- as.vector(classITS_tax_tab[classITS_tax_tab$Phylum=="Ascomycota", "Class"])
row.names(classITS_counts_tab) <- as.vector(classITS_tax_tab$Class)
#make table of Proteobacteria classes
Ascomycete_class_counts_tab <- classITS_counts_tab[row.names(classITS_counts_tab) %in% Ascomycete_classes_vec, ]
#make table ofProteobacteria not identified to the class level
Ascomycete_no_class_annotated_counts <- 
  phylaITS_and_unidentified_counts_tab[row.names(phylaITS_and_unidentified_counts_tab) %in% "Ascomycota",]-
  colSums(Ascomycete_class_counts_tab)
#Now combine the tables
majorITS_taxa_counts_tab <- rbind(temp_majorITS_taxa_counts_tab, Ascomycete_class_counts_tab,
                               "Unclassified_Ascomycota"=Ascomycete_no_class_annotated_counts)
head(majorITS_taxa_counts_tab)

```


```{r}
write.csv(majorITS_taxa_counts_tab, "majorITStaxa_count_rar_ITS_05.21.25.csv")
#Check that all sequences are accounted for
identical(colSums(majorITS_taxa_counts_tab), rowSums(asv_counts))
## [1] TRUE
#Convert totals to relative abundance
majorITS_taxa_proportions_tab <- apply(majorITS_taxa_counts_tab, 2, function(x) x/sum(x)*100)
#colSums(majorITS_taxa_proportions_tab)
write.csv(majorITS_taxa_proportions_tab, "majorITStaxa_relative abundance_rar_ITS_05.21.25.csv")

```


by family
```{r}
#make taxonomy object by genus
familyITS_counts_tab <- otu_table(tax_glom(pseqtestITS, taxrank="Family"), taxa_are_rows = FALSE)
familyITS_counts_tab <- t(familyITS_counts_tab)
#make vector of genus names to set as row names
familyITS_tax_vec <- as.vector(tax_table(tax_glom(pseqtestITS, taxrank="Family"))[,5]) 
rownames(familyITS_counts_tab) <- as.vector(familyITS_tax_vec)

#determine the number of unclassified seqs at the family level
unclassified_familyITS_counts <- colSums(t(asv_counts)) - colSums(familyITS_counts_tab)
unclassified_familyITS_counts
#Add a row of "unclassified" to the family count table
familyITS_and_unidentified_counts_tab <- rbind(familyITS_counts_tab, 
                                            "Unclassified_family"=unclassified_familyITS_counts)
write.csv(familyITS_and_unidentified_counts_tab, "familyITS raw abundance_rar_ITS_05.21.25.csv")
#Check that all seqs are accounted for.

identical(colSums(familyITS_and_unidentified_counts_tab), rowSums(asv_counts))

#Convert totals to relative abundance
familyITS_proportions_tab <- apply(familyITS_and_unidentified_counts_tab, 2, function(x) x/sum(x)*100)
write.csv(familyITS_proportions_tab, "family relative abundance_rar_ITS_05.21.25.csv")

```


```{r}
#Merge metadata, phylum, and family abundances
#phy_and_fam <- merge(t(majorITS_taxa_counts_tab), t(family_and_unidentified_counts_tab),
#                  by="row.names", all=TRUE)
phy_and_fam <- merge(t(majorITS_taxa_proportions_tab), t(familyITS_proportions_tab),
                  by="row.names", all=TRUE)
phyfam <- phy_and_fam[,-1]
rownames(phyfam) <- phy_and_fam[,1]
tax_merge <- merge(samdf_ITS, phyfam,
                  by="row.names", all=TRUE)
taxmergeITS <- tax_merge[,-1]
rownames(taxmergeITS) <- tax_merge[,1]
write.csv(tax_merge, "TALSYN_taxmergeITS_05.21.25.csv")

```

###Phylum-Class BARPLOT
```{r}
#majorITS taxa, mean percent abundances across all samples:
rowMeans(majorITS_taxa_proportions_tab)

#Select taxa that make up >90% of the total sequences

phylaITStrim <- data.frame(majorITS_taxa_proportions_tab[c("Basidiomycota","Mucoromycota","Mortierellomycota","Rozellomycota","Leotiomycetes","Dothideomycetes","Sordariomycetes","Eurotiomycetes","Lecanoromycetes","Unclassified_Ascomycota","Unclassified_Phylum"), ])

#Create an "other" category for the phyla not retained
filtered_proportions <- colSums(majorITS_taxa_proportions_tab) - 
  colSums(phylaITStrim)
phylaITStrim <- rbind(phylaITStrim, "Other"=filtered_proportions)
phylaITStrim

#Add taxa names as a column
phylaITStrim$Major_Taxa <- row.names(phylaITStrim)
#transform into long format
phylaITSlong <- gather(phylaITStrim, Sample, Proportion, -Major_Taxa)
phylaITSlong$Sample <- gsub("X","",phylaITSlong$Sample)

```

```{r}
## [1] "substrate" "WetDry"
phylaITSmet<-data.frame("Sample"=row.names(samdf_ITS),
                     "Substrate"=samdf_ITS$substrate,
                     "Classiification"=samdf_ITS$class_state,
                     stringsAsFactors=T)
#merge metadata with major taxa data
phylaITSlong <- merge(phylaITSlong, phylaITSmet)
#Summarize by depth and hydration
phylaITS_summary <- 
  phylaITSlong %>% # the names of the new data frame and the data frame to be summarised
  group_by(Substrate, Classiification, Major_Taxa) %>%   # the grouping variable
  summarise(mean_prop = mean(Proportion))  # calculates the mean of each group

phylaITS_summary

phylaITS_summary$Substrate <- factor(phylaITS_summary$Substrate,
                                 levels=c("L","B","S")) #reorder variables

#phylaITS_summary$Wet.Dry <- factor(phylaITS_summary$Wet.Dry, levels= c("wet","dry")) #reorder variables
phylaITS_summary$Major_Taxa <- factor(phylaITS_summary$Major_Taxa, levels= rev(c("Basidiomycota","Mucoromycota","Mortierellomycota","Rozellomycota","Leotiomycetes","Dothideomycetes","Sordariomycetes","Eurotiomycetes","Lecanoromycetes","Unclassified_Ascomycota","Unclassified_Phylum","Other")))

#color palette
sublab<- c("Leaf litter","Epilithic biofilms","Benthic sediment")
names(sublab)<- c("L","B","S")

pal12 <- rev(cols25(n=12)) #set n= to the number of taxa you are plotting
#make stacked bar plot
phylumITS_bar <- ggplot(phylaITS_summary, aes(x = Classiification, y = mean_prop, fill = Major_Taxa))+
  geom_bar(stat = "identity", col=I("black")) +
  scale_fill_manual(values=pal12)+
  guides(fill=guide_legend(ncol=1))+
  facet_wrap(~Substrate, labeller = labeller(Substrate=sublab), nrow=1, scales="free_x") +
  labs(x=NULL,y="Relative abundance (%)",
       fill="Major taxa")+
  #scale_facet_discrete(labels=c('B'='Epilithic biofilms', 'L'='Leaf litter', 'S'='Benthic sediment', 'W'='Surface water'))+
  theme(axis.text.x = element_text(angle=60, hjust=1), legend.position = "right")+
  theme(text=element_text(size=18), #change font size of all text
        axis.text=element_text(size=16), #change font size of axis text
        axis.title=element_text(size=16), #change font size of axis titles
        plot.title=element_text(size=16), #change font size of plot title
        legend.text=element_text(size=16), #change font size of legend text
        strip.text.x = element_text(size = 16),
        legend.title=element_text(size=16)) #change font size of legend title  
phylumITS_bar
 
#c(180,100) *
#  0.0394 * # convert mm to inch
#  600 # convert to pixels
## [1] 4255.2 2364.0
plotout <- "phylumITS_barplot__052125.tiff"
agg_tiff(filename=plotout, width=4255, height=2364, units="px",
         pointsize=10, res=600, compression="lzw", scaling=0.5)
phylumITS_bar
invisible(dev.off())
```
```{r}
library(cowplot)
spacer<- ggdraw()
plotout <- "taxa-2-panel_05-22-2025.tiff"
agg_tiff(filename=plotout, width=4255, height=4728, units="px",
         pointsize=10, res=600, compression="lzw", scaling=0.5)
plot_grid(phylum16S_bar,spacer, phylumITS_bar,rel_heights = c(1,0.05,1) , labels=c('A','','B'),label_size = 18, ncol = 1)
invisible(dev.off())
```


##

```{r}
library(rstatix)
stat.test <- kingmerge %>%
wilcox_test(Archaea~substrate) %>%
  adjust_pvalue(method = "bonferroni") %>%
  add_significance()
stat.test
effsize <- kingmerge %>%
  wilcox_effsize(Archaea~substrate)
effsize

cor_summary <- taxmerge16S[!is.na(taxmerge16S$percentwet_11month),] %>%
  group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Methanobacteriaceae, percentwet_11month, method = "spearman"),
    p_value = cor.test(Methanobacteriaceae, percentwet_11month, method = "spearman")$p.value
  )
cor_summary

cor_summary <- taxmerge16S[!is.na(taxmerge16S$CH4_avg),] %>%
  group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Methanobacteriaceae, CH4_avg, method = "spearman"),
    p_value = cor.test(Methanobacteriaceae, CH4_avg, method = "spearman")$p.value
  )
cor_summary

```

## Class merge
#### Prokaryote phyla and classes

```{r}
#make taxonomy object by phylum
class16S_counts_tab <- otu_table(tax_glom(pseqtest16S, taxrank="Class"), taxa_are_rows = FALSE)
class16S_counts_tab <- t(class16S_counts_tab)
#make vector of phyla names to set as row names
class16S_counts_vec <- as.vector(tax_table(tax_glom(pseqtest16S, taxrank="Class"))[,3]) 
rownames(class16S_counts_tab) <- as.vector(class16S_counts_vec)
class16S_counts_tab
write.csv(class16S_counts_tab, 'class16S_counts_05.28.25.csv')
          
```

```{r}
asv_counts <- pseqtest16S@otu_table
## determine the number of unclassified seqs at the class level
unclassified_tax_counts <- colSums(t(asv_counts)) - colSums(class16S_counts_tab)
## Add a row of "unclassified" to the phylum count table
class16S_and_unidentified_counts_tab <- rbind(class16S_counts_tab, "Unclassified_Class"=unclassified_tax_counts)

```

```{r}
#Check that all sequences are accounted for
identical(colSums(class16S_and_unidentified_counts_tab), rowSums(asv_counts))
## [1] TRUE
#Convert totals to relative abundance
class16S_proportions_tab <- apply(class16S_and_unidentified_counts_tab, 2, function(x) x/sum(x)*100)
write.csv(class16S_proportions_tab, "class16S_proportions_tab_05.21.25.csv")

```

```{r}
class_merge16S <- merge(samdf_16S, t(class16S_proportions_tab),
                  by="row.names", all=TRUE)
rownames(class_merge16S) <- class_merge16S[,1]
class_merge16S <- class_merge16S[,-1]

write.csv(class_merge16S, "TALSYN_classmerge16S_05.28.25.csv")

rowMeans(class16S_proportions_tab)

```
Cyanobacteriia
```{r}
library(rstatix)

cor_summary <- class_merge16S[!is.na(class_merge16S$percentwet_11month)&class_merge16S$substrate=='B',] %>%
  #group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Cyanobacteriia, percentwet_11month, method = "spearman"),
    p_value = cor.test(Cyanobacteriia, percentwet_11month, method = "spearman")$p.value
  )
cor_summary
cor_summary <- class_merge16S[!is.na(class_merge16S$percentwet_11month)&class_merge16S$substrate=='B',] %>%
  #group_by(substrate) %>%
  summarize(
    spearman_rho = cor(B_Chla, percentwet_11month, method = "spearman"),
    p_value = cor.test(B_Chla, percentwet_11month, method = "spearman")$p.value
  )
cor_summary

cor_summary <- class_merge16S[!is.na(class_merge16S$percentwet_11month)&!is.na(class_merge16S$W_Chla)&class_merge16S$substrate=='L',] %>%
  #group_by(substrate) %>%
  summarize(
    spearman_rho = cor(W_Chla, percentwet_11month, method = "spearman"),
    p_value = cor.test(W_Chla, percentwet_11month, method = "spearman")$p.value
  )
cor_summary

```


Based on https://doi.org/10.1128/aem.02247-23 
Known methanogenic classes of Archaea include Thermoplasmia (Thermoplasmata), Methanobacteriia (Methanobacteria), Methanocellia, Methanonatronarchaeia, Methanomicrobia, Metanephric, and Methanococci 

```{r}
library(rstatix)

## Methanogenic Archaea classes present in our study
# Methanobacteria; Methanomicrobia; Thermoplasmata; Methanocellia; Methanomethylicia; Methanosarcinia

## non-methanogeenix archaea: Nitrososphaeria

## Methanobacteria
cor_summary <- class_merge16S[!is.na(class_merge16S$percentwet_11month),] %>%
  group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Methanobacteria, percentwet_11month, method = "spearman"),
    p_value = 3*cor.test(Methanobacteria, percentwet_11month, method = "spearman")$p.value
  )
cor_summary
cor_summary <- class_merge16S[!is.na(class_merge16S$CH4_avg),] %>%
  group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Methanobacteria, CH4_avg, method = "spearman"),
    p_value = 3*cor.test(Methanobacteria, CH4_avg, method = "spearman")$p.value
  )
cor_summary

## Methanomicrobia
cor_summary <- class_merge16S[!is.na(class_merge16S$percentwet_11month),] %>%
  group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Methanomicrobia, percentwet_11month, method = "spearman"),
    p_value = 3*cor.test(Methanomicrobia, percentwet_11month, method = "spearman")$p.value
  )
cor_summary
cor_summary <- class_merge16S[!is.na(class_merge16S$CH4_avg),] %>%
  group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Methanomicrobia, CH4_avg, method = "spearman"),
    p_value = 3*cor.test(Methanomicrobia, CH4_avg, method = "spearman")$p.value
  )
cor_summary

## Thermoplasmata
cor_summary <- class_merge16S[!is.na(class_merge16S$percentwet_11month),] %>%
  group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Thermoplasmata, percentwet_11month, method = "spearman"),
    p_value = 3*cor.test(Thermoplasmata, percentwet_11month, method = "spearman")$p.value
  )
cor_summary
cor_summary <- class_merge16S[!is.na(class_merge16S$CH4_avg),] %>%
  group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Thermoplasmata, CH4_avg, method = "spearman"),
    p_value = 3*cor.test(Thermoplasmata, CH4_avg, method = "spearman")$p.value
  )
cor_summary

## Methanocellia
cor_summary <- class_merge16S[!is.na(class_merge16S$percentwet_11month),] %>%
  group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Methanocellia, percentwet_11month, method = "spearman"),
    p_value = 3*cor.test(Methanocellia, percentwet_11month, method = "spearman")$p.value
  )
cor_summary
cor_summary <- class_merge16S[!is.na(class_merge16S$CH4_avg),] %>%
  group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Methanocellia, CH4_avg, method = "spearman"),
    p_value = 3*cor.test(Methanocellia, CH4_avg, method = "spearman")$p.value
  )
cor_summary

## Methanomethylicia
cor_summary <- class_merge16S[!is.na(class_merge16S$percentwet_11month),] %>%
  group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Methanomethylicia, percentwet_11month, method = "spearman"),
    p_value = 3*cor.test(Methanomethylicia, percentwet_11month, method = "spearman")$p.value
  )
cor_summary
cor_summary <- class_merge16S[!is.na(class_merge16S$CH4_avg),] %>%
  group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Methanomethylicia, CH4_avg, method = "spearman"),
    p_value = 3*cor.test(Methanomethylicia, CH4_avg, method = "spearman")$p.value
  )
cor_summary

## Methanosarcinia
cor_summary <- class_merge16S[!is.na(class_merge16S$percentwet_11month),] %>%
  group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Methanosarcinia, percentwet_11month, method = "spearman"),
    p_value = 3*cor.test(Methanosarcinia, percentwet_11month, method = "spearman")$p.value
  )
cor_summary
cor_summary <- class_merge16S[!is.na(class_merge16S$CH4_avg),] %>%
  group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Methanosarcinia, CH4_avg, method = "spearman"),
    p_value = 3*cor.test(Methanosarcinia, CH4_avg, method = "spearman")$p.value
  )
cor_summary


###Nitrososphaeria
cor_summary <- class_merge16S[!is.na(class_merge16S$percentwet_11month),] %>%
  group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Nitrososphaeria, percentwet_11month, method = "spearman"),
    p_value = cor.test(Nitrososphaeria, percentwet_11month, method = "spearman")$p.value
  )
cor_summary

cor_summary <- class_merge16S[!is.na(class_merge16S$CH4_avg),] %>%
  group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Nitrososphaeria, CH4_avg, method = "spearman"),
    p_value = cor.test(Nitrososphaeria, CH4_avg, method = "spearman")$p.value
  )
cor_summary

cor_summary <- class_merge16S[!is.na(class_merge16S$N2Ar_avg),] %>%
  group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Nitrososphaeria, N2Ar_avg, method = "spearman"),
    p_value = cor.test(Nitrososphaeria, N2Ar_avg, method = "spearman")$p.value
  )
cor_summary

cor_summary <- class_merge16S[!is.na(class_merge16S$N2O_avg),] %>%
  group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Nitrososphaeria, N2O_avg, method = "spearman"),
    p_value = cor.test(Nitrososphaeria, N2O_avg, method = "spearman")$p.value
  )
cor_summary

cor_summary <- class_merge16S[!is.na(class_merge16S$CO2_avg),] %>%
  group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Nitrososphaeria, CO2_avg, method = "spearman"),
    p_value = cor.test(Nitrososphaeria, CO2_avg, method = "spearman")$p.value
  )
cor_summary

cor_summary <- class_merge16S[!is.na(class_merge16S$O2_uM_avg),] %>%
  group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Nitrososphaeria, O2_uM_avg, method = "spearman"),
    p_value = cor.test(Nitrososphaeria, O2_uM_avg, method = "spearman")$p.value
  )
cor_summary
```

Archaeal class in sediment only, with bonferonni correction
```{r}
class_merge16S_S<- class_merge16S[class_merge16S$substrate=='S',]

## Methanobacteria
cor_summary <- class_merge16S_S[!is.na(class_merge16S_S$percentwet_11month),] %>%
 # group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Methanobacteria, percentwet_11month, method = "spearman"),
    p_value = 7*cor.test(Methanobacteria, percentwet_11month, method = "spearman")$p.value
  )
cor_summary

## Methanomicrobia
cor_summary <- class_merge16S_S[!is.na(class_merge16S_S$percentwet_11month),] %>%
 # group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Methanomicrobia, percentwet_11month, method = "spearman"),
    p_value = 7*cor.test(Methanomicrobia, percentwet_11month, method = "spearman")$p.value
  )
cor_summary

## Thermoplasmata
cor_summary <- class_merge16S_S[!is.na(class_merge16S_S$percentwet_11month),] %>%
 # group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Thermoplasmata, percentwet_11month, method = "spearman"),
    p_value = 7*cor.test(Thermoplasmata, percentwet_11month, method = "spearman")$p.value
  )
cor_summary


## Methanocellia
cor_summary <- class_merge16S_S[!is.na(class_merge16S_S$percentwet_11month),] %>%
 # group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Methanocellia, percentwet_11month, method = "spearman"),
    p_value = 7*cor.test(Methanocellia, percentwet_11month, method = "spearman")$p.value
  )
cor_summary

## Methanomethylicia
cor_summary <- class_merge16S_S[!is.na(class_merge16S_S$percentwet_11month),] %>%
 # group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Methanomethylicia, percentwet_11month, method = "spearman"),
    p_value = 7*cor.test(Methanomethylicia, percentwet_11month, method = "spearman")$p.value
  )
cor_summary


## Methanosarcinia
cor_summary <- class_merge16S_S[!is.na(class_merge16S_S$percentwet_11month),] %>%
 # group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Methanosarcinia, percentwet_11month, method = "spearman"),
    p_value = 7*cor.test(Methanosarcinia, percentwet_11month, method = "spearman")$p.value
  )
cor_summary

###Nitrososphaeria
cor_summary <- class_merge16S_S[!is.na(class_merge16S_S$percentwet_11month),] %>%
 # group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Nitrososphaeria, percentwet_11month, method = "spearman"),
    p_value = 7*cor.test(Nitrososphaeria, percentwet_11month, method = "spearman")$p.value
  )
cor_summary

##### METHANE #######################
## Methanobacteria
cor_summary <- class_merge16S_S[!is.na(class_merge16S_S$CH4_avg),] %>%
 # group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Methanobacteria, CH4_avg, method = "spearman"),
    p_value = 7*cor.test(Methanobacteria, CH4_avg, method = "spearman")$p.value
  )
cor_summary

## Methanomicrobia
cor_summary <- class_merge16S_S[!is.na(class_merge16S_S$CH4_avg),] %>%
 # group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Methanomicrobia, CH4_avg, method = "spearman"),
    p_value = 7*cor.test(Methanomicrobia, CH4_avg, method = "spearman")$p.value
  )
cor_summary

## Thermoplasmata
cor_summary <- class_merge16S_S[!is.na(class_merge16S_S$CH4_avg),] %>%
 # group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Thermoplasmata, CH4_avg, method = "spearman"),
    p_value = 7*cor.test(Thermoplasmata, CH4_avg, method = "spearman")$p.value
  )
cor_summary


## Methanocellia
cor_summary <- class_merge16S_S[!is.na(class_merge16S_S$CH4_avg),] %>%
 # group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Methanocellia, CH4_avg, method = "spearman"),
    p_value = 7*cor.test(Methanocellia, CH4_avg, method = "spearman")$p.value
  )
cor_summary

## Methanomethylicia
cor_summary <- class_merge16S_S[!is.na(class_merge16S_S$CH4_avg),] %>%
 # group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Methanomethylicia, CH4_avg, method = "spearman"),
    p_value = 7*cor.test(Methanomethylicia, CH4_avg, method = "spearman")$p.value
  )
cor_summary


## Methanosarcinia
cor_summary <- class_merge16S_S[!is.na(class_merge16S_S$CH4_avg),] %>%
 # group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Methanosarcinia, CH4_avg, method = "spearman"),
    p_value = 7*cor.test(Methanosarcinia, CH4_avg, method = "spearman")$p.value
  )
cor_summary

###Nitrososphaeria
cor_summary <- class_merge16S_S[!is.na(class_merge16S_S$CH4_avg),] %>%
 # group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Nitrososphaeria, CH4_avg, method = "spearman"),
    p_value = 7*cor.test(Nitrososphaeria, CH4_avg, method = "spearman")$p.value
  )
cor_summary

```

```{r}
pseqtest16Sarch<- subset_taxa(pseqtest16S, Kingdom=="Archaea")
#make taxonomy object by phylum
class16Sarch_counts_tab <- otu_table(tax_glom(pseqtest16Sarch, taxrank="Class"), taxa_are_rows = FALSE)
class16Sarch_counts_tab <- t(class16Sarch_counts_tab)
#make vector of phyla names to set as row names
class16Sarch_counts_vec <- as.vector(tax_table(tax_glom(pseqtest16Sarch, taxrank="Class"))[,3]) 
rownames(class16Sarch_counts_tab) <- as.vector(class16Sarch_counts_vec)
class16Sarch_counts_tab
write.csv(class16Sarch_counts_tab, 'class16Sarch_counts_05.28.25.csv')
          
```

```{r}
asv_counts <- pseqtest16Sarch@otu_table
## determine the number of unclassified seqs at the class level
unclassified_tax_counts <- colSums(t(asv_counts)) - colSums(class16Sarch_counts_tab)
## Add a row of "unclassified" to the phylum count table
class16Sarch_and_unidentified_counts_tab <- rbind(class16Sarch_counts_tab, "Unclassified_Class"=unclassified_tax_counts)

```

```{r}
#Check that all sequences are accounted for
identical(colSums(class16Sarch_and_unidentified_counts_tab), rowSums(asv_counts))
## [1] TRUE
#Convert totals to relative abundance
class16Sarch_proportions_tab <- apply(class16Sarch_and_unidentified_counts_tab, 2, function(x) x/sum(x)*100)
write.csv(class16Sarch_proportions_tab, "class16Sarch_proportions_tab_05.21.25.csv")

class16Sarch_proportions_tab<-class16Sarch_proportions_tab[,!is.na(colSums(class16Sarch_proportions_tab)>0)]
#class16Sarch_proportions_tab<-class16Sarch_proportions_tab[,!is.na(colSums(class16Sarch_proportions_tab))]
rowMeans(class16Sarch_proportions_tab)

```

```{r}
samdf_16Sarch<- sam_data(pseqtest16Sarch)
class_merge16Sarch <- merge(samdf_16Sarch, t(class16Sarch_proportions_tab),
                  by="row.names", all=TRUE)
rownames(class_merge16Sarch) <- class_merge16Sarch[,1]
class_merge16Sarch <- class_merge16Sarch[,-1]

write.csv(class_merge16Sarch, "TALSYN_classmerge16Sarch_05.28.25.csv")
## detatch linkET package, it interferes with downstream stats

```

Ephemeral sediment Basidiomycota:
taxmergeITS
```{r}
taxmergeITS_S<- taxmergeITS[taxmergeITS$substrate=='S',]

taxmergeITS_Seph<- taxmergeITS_S[taxmergeITS_S$int_eph=='ephemeral',]

mean(taxmergeITS_Seph$Basidiomycota)
sd(taxmergeITS_Seph$Basidiomycota)

stat.test <- taxmergeITS_S %>%
wilcox_test(Basidiomycota~int_eph) %>%
  adjust_pvalue(method = "bonferroni") %>%
  add_significance()
stat.test
effsize <-  taxmergeITS_S %>%
  wilcox_effsize(Basidiomycota~int_eph)
effsize


## Basidiomycota
cor_summary <- taxmergeITS_S[!is.na(taxmergeITS_S$percentwet_11month),] %>%
 # group_by(substrate) %>%
  summarize(
    spearman_rho = cor(Basidiomycota, percentwet_11month, method = "spearman"),
    p_value = cor.test(Basidiomycota, percentwet_11month, method = "spearman")$p.value
  )
cor_summary


```
