---
title: "Analysis_0.5_functional_traits"
author: "Charles T. Bond"
date: "`r Sys.Date()`"
output: html_document
---

# Microbial Community Analysis: Talladega Forest Non-perennial Stream Network

## Analysis_0.5: Functional (PICRUSt2 predicted gene relative abundance) analysis

### Introduction
This code(0.5) will generate figures and tests to look for patterns in PICRUSt2-based functional traits (predicted relative abunance of functionally important genes). First we will use the MetaCyc outputs of PICRUSt2 to look for trends in broad functional pathways (photosynthesis, methanogenesis, and various aerobic and anaerobic heterotrophic pathways). Then we wil use KEGG orthology for specific genes aassociatedd with different steps in denitrification. 


### Setup
```{r, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/chunk/Talladega_synoptic")
knitr::opts_knit$set(root.dir = "/Users/chunk/Talladega_synoptic/Analysis")
options(knitr.duplicate.label = "allow") ### so I don't have to give chunks unique names
```

```{r}
### loading libraries
library(readr)
library(ggpicrust2)
library(tibble)
library(tidyverse)
library(ggprism)
library(patchwork)
### loading metadata
metadata_full <- read.csv("meta_sub_05.12.2025.csv", row.names=1)
metadata_full$substrate <- as.factor(metadata_full$substrate)

```


### METACYC
##### Selected pathways representing diversion metabolisms
Photoautotrophy (sucrose synthesis): SUCSYN-PWY        
Bacteriochlorophyll biosynthesis (Photoautotrophy): PWY-5531    
Chlorophyllide-a biosynthesis (Photoautotrophy):  PWY-7159
Chlorophyll synthesis: CHLOROPHYLL-SYN
Complete reductive TCA (autotrophy): P23-PWY
Incomplete reductive TCA (methanogenic autotrophy): P42-PWY
Methanogenesis: METHANOGENESIS-PWY
Methyltrophy: PWY-1882   
Glycolysis: GLYCOLYSIS
Glucose and Xylose Degradation: PWY-6901  
Heterolactic Fermentation: P122-PWY    
Homolactic Fermentation: 	ANAEROFRUCAT-PWY
Denitrification (overall): DENITRIFICATION-PWY
Sulfur oxidation: PWY-5304


```{r}
setwd('/Users/chunk/Talladega_synoptic/picrust2')

#load pathway counts and ASV counts
pathway_abundance <- as.data.frame(read_tsv("/Users/chunk/Talladega_synoptic/Analysis/picrust2/picrust2_out_pipeline/pathways_out/path_abun_unstrat.tsv"))
row.names(pathway_abundance)<- pathway_abundance[,1]
pathway_abundance<-pathway_abundance[,-1]
pathway_abundance<-as.data.frame(t(pathway_abundance))

### extract ecologically relevant metabolic groups
eco_path <- pathway_abundance[,c("SUCSYN-PWY",
                                 "PWY-5531",
                                 "PWY-7159",
                                 "CHLOROPHYLL-SYN",
                                 "P23-PWY",
                                 "P42-PWY",
                                 "METHANOGENESIS-PWY",
                                 "PWY-1882",
                                 "GLYCOLYSIS",
                                 "PWY-6901",
                                 "P122-PWY",
                                 "ANAEROFRUCAT-PWY",
                                 "DENITRIFICATION-PWY",
                                 "PWY-5304")]

ecotab<- merge(metadata_full, eco_path, by= 'row.names')
row.names(ecotab)<- ecotab[,1]
ecotab<-ecotab[,-1]

### removing the beaver pond as an outlier
ecotab<- ecotab[ecotab$siteId!='TLMBP',]

## we want full 11 month percent wet data
ecotab<- ecotab[!is.na(ecotab$percentwet_11month),]
ecotab_S<- ecotab[ecotab$substrate=='S',]
ecotab_L<- ecotab[ecotab$substrate=='L',]
ecotab_B<- ecotab[ecotab$substrate=='B',]

```

### MetaCyc pathways GLLVM by microhabitat
Sediment
```{r}
library(stringr)
library(mvabund)
library(gllvm)
library(tidyverse)

#detach("package:linkET", unload=TRUE)
eco_abun<-ecotab_S[,101:114]
X <- as.matrix(ecotab_S[,c("percentwet_11month","LL_AFDM_prc")])
y<-as.matrix(as.tibble(eco_abun))
colnames(y)
## impoving interperetability of colnames for pathways with only numbered label
pwyname<- c("Sucrose synthesis (SUCSYN-PWY)", "Bacteriochlorophyll biosynthesis (PWY-5531)", "Chlorophyllide-a biosynthesis (PWY-7159)", "CHLOROPHYLL-SYN", "Complete reductive TCA (P23-PWY)", "Incomplete reductive TCA (P42-PWY)", "METHANOGENESIS-PWY", "Methyltrophy (PWY-1882)","GLYCOLYSIS","Glucose and Xylose Degradation (PWY-6901)", "Heterolactic Fermentation (P122-PWY)", "Homolactic Fermentation (ANAEROFRUCAT-PWY)", "DENITRIFICATION-PWY", "Sulfur oxidation (PWY-5304)")
colnames(y)<-pwyname

gllvm(y, family = "poisson")
gllvm(y, family = "negative.binomial")
fit_ord<- gllvm(y, family = "negative.binomial")
ordiplot(fit_ord, biplot=TRUE, ind.spp=14)

#gllvm(y, X, family = "negative.binomial")
#gllvm(y, family = "negative.binomial")
#gllvm(formula = y~X$prc_wet, family = "negative.binomial")

fit_env <- gllvm(y, X, family = "negative.binomial", num.lv = 1,
                 formula = ~ percentwet_11month, seed = 1234)
coefplot(fit_env, cex.ylab = 0.7, mar = c(4, 9, 2, 1), mfrow=c(1,1))

pdf("GLLVM_MetaCyc_S_plot.pdf", 
    )
gllvm_S<- coefplot(fit_env, cex.ylab = 0.7, mar = c(4, 15, 2, 1), mfrow=c(1,1))
gllvm_S
dev.off()
```

Leaf
```{r}
library(stringr)
library(mvabund)
library(gllvm)
library(tidyverse)

#detach("package:linkET", unload=TRUE)
eco_abun<-ecotab_L[,101:114]
X <- as.matrix(ecotab_L[,c("percentwet_11month","LL_AFDM_prc")])
y<-as.matrix(as.tibble(eco_abun))
colnames(y)
## impoving interperetability of colnames for pathways with only numbered label
pwyname<- c("Sucrose synthesis (SUCSYN-PWY)", "Bacteriochlorophyll biosynthesis (PWY-5531)", "Chlorophyllide-a biosynthesis (PWY-7159)", "CHLOROPHYLL-SYN", "Complete reductive TCA (P23-PWY)", "Incomplete reductive TCA (P42-PWY)", "METHANOGENESIS-PWY", "Methyltrophy (PWY-1882)","GLYCOLYSIS","Glucose and Xylose Degradation (PWY-6901)", "Heterolactic Fermentation (P122-PWY)", "Homolactic Fermentation (ANAEROFRUCAT-PWY)", "DENITRIFICATION-PWY", "Sulfur oxidation (PWY-5304)")
colnames(y)<-pwyname

gllvm(y, family = "poisson")
gllvm(y, family = "negative.binomial")
fit_ord<- gllvm(y, family = "negative.binomial")
ordiplot(fit_ord, biplot=TRUE, ind.spp=14)

#gllvm(y, X, family = "negative.binomial")
#gllvm(y, family = "negative.binomial")
#gllvm(formula = y~X$prc_wet, family = "negative.binomial")

fit_env <- gllvm(y, X, family = "negative.binomial", num.lv = 1,
                 formula = ~ percentwet_11month, seed = 1234)
coefplot(fit_env, cex.ylab = 0.7, mar = c(4, 9, 2, 1), mfrow=c(1,1))

pdf("GLLVM_MetaCyc_L_plot.pdf", 
    )
gllvm_L<- coefplot(fit_env, cex.ylab = 0.7, mar = c(4, 15, 2, 1), mfrow=c(1,1))
gllvm_L
dev.off()
```
 
Rock
```{r}
library(stringr)
library(mvabund)
library(gllvm)
library(tidyverse)

#detach("package:linkET", unload=TRUE)
eco_abun<-ecotab_B[,101:114]
X <- as.matrix(ecotab_B[,c("percentwet_11month","LL_AFDM_prc")])
y<-as.matrix(as.tibble(eco_abun))
colnames(y)
## impoving interperetability of colnames for pathways with only numbered label
pwyname<- c("Sucrose synthesis (SUCSYN-PWY)", "Bacteriochlorophyll biosynthesis (PWY-5531)", "Chlorophyllide-a biosynthesis (PWY-7159)", "CHLOROPHYLL-SYN", "Complete reductive TCA (P23-PWY)", "Incomplete reductive TCA (P42-PWY)", "METHANOGENESIS-PWY", "Methyltrophy (PWY-1882)","GLYCOLYSIS","Glucose and Xylose Degradation (PWY-6901)", "Heterolactic Fermentation (P122-PWY)", "Homolactic Fermentation (ANAEROFRUCAT-PWY)", "DENITRIFICATION-PWY", "Sulfur oxidation (PWY-5304)")
colnames(y)<-pwyname

gllvm(y, family = "poisson")
gllvm(y, family = "negative.binomial")
fit_ord<- gllvm(y, family = "negative.binomial")
ordiplot(fit_ord, biplot=TRUE, ind.spp=14)

#gllvm(y, X, family = "negative.binomial")
#gllvm(y, family = "negative.binomial")
#gllvm(formula = y~X$prc_wet, family = "negative.binomial")

fit_env <- gllvm(y, X, family = "negative.binomial", num.lv = 1,
                 formula = ~ percentwet_11month, seed = 1234)
coefplot(fit_env, cex.ylab = 0.7, mar = c(4, 9, 2, 1), mfrow=c(1,1))

pdf("GLLVM_MetaCyc_B_plot.pdf", 
    )
gllvm_B<- coefplot(fit_env, cex.ylab = 0.7, mar = c(4, 15, 2, 1), mfrow=c(1,1))
gllvm_B
dev.off()
```
 

 
 ########################################################
 

# KEGG Orthology for more in-depth look at denitrification. 
## The twelve KO identifiers for denitrification genes:
The 12 genes and associated KEGG Orthology Identifiers are narG (K00370), narH (K00371), narI (K00374), napA (K02567), napB (K02568), nirK (K00368), nirS (K15864), norB (K04561), norC (K02305), norV (K12264), norW (K12265), and nosZ (K00376). 
More details from Isokpehi eet al. https://doi.org/10.3390/microorganisms12040791 
K00370	nitrate reductase/nitrite oxidoreductase, alpha subunit	narG
K00371	nitrate reductase/nitrite oxidoreductase, beta subunit	narH
K00374	nitrate reductase gamma subunit	narI
K02567	nitrate reductase (cytochrome)	napA
K02568	nitrate reductase (cytochrome), electron transfer subunit	napB
K00368	nitrite reductase (NO-forming)	nirK
K15864	nitrite reductase (NO-forming)/hydroxylamine reductase	nirS
K04561	nitric oxide reductase subunit B	norB
K02305	nitric oxide reductase subunit C	norC
K12264	anaerobic nitric oxide reductase flavorubredoxin	norV
K12265	nitric oxide reductase FlRd-NAD(+) reductase	norW
K00376	nitrous-oxide reductase	nosZ
```{r}
setwd('/Users/chunk/Talladega_synoptic/picrust2')

#load pathway counts and ASV counts
pathway_abundance <- as.data.frame(read_tsv("/Users/chunk/Talladega_synoptic/Analysis/picrust2/picrust2_out_pipeline/KO_metagenome_out/pred_metagenome_unstrat.tsv"))

rownames(pathway_abundance)<-pathway_abundance$`function`
pathway_abundance<-pathway_abundance[,-1]
pathway_abundance<- as.data.frame(t(pathway_abundance))

keggabun<- cbind("narG (K00370)"=pathway_abundance$K00370,
                                "narH (K00371)"=pathway_abundance$K00371,
                                "narI (K00374)"=pathway_abundance$K00374,
                                "napA (K02567)"=pathway_abundance$K02567,
                                "napB (K02568)"=pathway_abundance$K02568,
                                "nirK (K00368)"=pathway_abundance$K00368,
                                "nirS (K15864)"=pathway_abundance$K15864,
                                "norB (K04561)"=pathway_abundance$K04561,
                                "norC (K02305)"=pathway_abundance$K02305,
                                "norV (K12264)"=pathway_abundance$K12264,
                                "norW (K12265)"=pathway_abundance$K12265,
                                "nosZ (K00376)"=pathway_abundance$K00376)
rownames(keggabun)<-rownames(pathway_abundance)

ecotab<- merge(metadata_full, keggabun, by= 'row.names')
row.names(ecotab)<- ecotab$Row.names
ecotab$substrate<- as.factor(ecotab$substrate)
ecotab[,102:113]<- ecotab[,102:113] %>% mutate_if(is.character,as.numeric)

```

```{r}
ecotab[,102:113]
ecotabS<- ecotab[ecotab$substrate=="S"&!is.na(ecotab$percentwet_11month),]

library(linkET)
correlate(ecotabS[,102:113], ecotabS[,c(82:83,88:89)]) %>% 
  qcorrplot() +
  geom_square() +
    geom_mark(sep = '\n',size=4,sig_level = c(0.05,0.01,0.001),sig_thres = 0.05,
            only_mark = TRUE)+
  scale_fill_gradientn(colours = RColorBrewer::brewer.pal(11, "RdBu"))+
  xlab(label="")+
  ylab(label="")
```
```{r}
library(linkET)
ecotabS<- ecotab[ecotab$substrate=="S"&!is.na(ecotab$percentwet_11month)&!is.na(ecotab$CH4_avg),]

adjust_pvalue(correlate(ecotabS[,102:113], ecotabS[,c(65,75,89)]),method = "holm") %>% 
  qcorrplot() +
  geom_square() +
    geom_mark(sep = '\n',size=4,sig_level = c(0.05,0.01,0.001),sig_thres = 0.05,
            only_mark = TRUE)+
  scale_fill_gradientn(colours = RColorBrewer::brewer.pal(11, "RdBu"))+
  xlab(label="")+
  ylab(label="")+
  ggtitle(label="Sediments (n=31)")
```





