---
title: "Analysis_0.2_Community_Composition_Analysis"
author: "Charles T. Bond"
date: "`r Sys.Date()`"
output: html_document
---

# Microbial Community Analysis: Talladega Forest Non-perennial Stream Network

## Analysis_0.2: Community Composition Across a Flow-Permanence Continuum

### Introduction
In the previous R markdown, we processed 16S and ITS ASV count data, visualized, and rarefied  Now the rarefied 16S and ITS data, as well as accumulation curves and Venn Diagrams, are exported at .Rdata files. 
This is helpful as my laptop's RAM is a bit overwhelmed handling the full 16S data, so being able to start fresh with just the rarefied data should help things run a little smoother. 

## Setup
```{r, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/chunk/Talladega_synoptic")
knitr::opts_knit$set(root.dir = "/Users/chunk/Talladega_synoptic/Analysis")
options(knitr.duplicate.label = "allow") ### so I don't have to give chunks unique names
```

```{r}
#library(dada2)
library(phyloseq)
library(vegan)
library(ggplot2)
library(tidyr)
library(dplyr)
library(pheatmap)
library(GUniFrac)
library(pals)
library(RColorBrewer)
library(ragg)
library(ggpubr)

#Set theme
theme_set(theme_bw() + theme(
              plot.title = element_text(size=20, color="black"),
              axis.text.x = element_text(size=15, color="black"),
              axis.text.y = element_text(size=15, color="black"),
              axis.title.x = element_text(size=15),
              axis.title.y = element_text(size=15),
              legend.text = element_text(size=12),
              legend.title = element_text(size=15),
            #  legend.position = "bottom",
            #  legend.key=element_blank(),
            #  legend.key.size = unit(0.5, "cm"),
            #  legend.spacing.x = unit(0.1, "cm"),
            #  legend.spacing.y = unit(0.1, "cm"),
              panel.background = element_blank(), 
              #panel.border = element_rect(colour = "black", fill=NA, size=1),
              plot.background = element_blank()))
```

## Load phyloseq objects
```{r}
#prokaryotes
load("/Users/chunk/Talladega_synoptic/Analysis/Rdata/pseqtest16S.rdata")
pseqtest16S

#fungi
load("/Users/chunk/Talladega_synoptic/Analysis/Rdata/pseqtestITS.rdata")
pseqtestITS

## Cutting the single beaver pond site, its an outlier, the only permanent lentic site
pseqtest16S <- subset_samples(pseqtest16S, siteId != "TLMBP")
pseqtestITS <- subset_samples(pseqtestITS, siteId != "TLMBP")

# extract sampe data
samdf_16S <- data.frame(sample_data(pseqtest16S))
samdf_ITS <- data.frame(sample_data(pseqtestITS))

```

## 0.2.1 
### Bray-Curtis NMDS

#### 16S
```{r}
#generate distance matrices
bray_16S <- phyloseq::distance(pseqtest16S, method = "bray")
#jactest <- phyloseq::distance(pseqtest16S, method="jaccard", binary=TRUE)
# NMDS
braymds_16S <- metaMDS(bray_16S, k=2, trymax=500, wascores = T)
brayscores_16S <- scores(braymds_16S)
#jacmds <- metaMDS(jactest, k=3, trymax=500, wascores = T)
#jacscores <- scores(jacmds)
braymds_16S
#jacmds

stressplot(braymds_16S)
#stressplot(jacmds)

factorfit(brayscores_16S, samdf_16S$substrate, permutations = 999)
factorfit(brayscores_16S, samdf_16S$WetDry, strata = samdf_16S$substrate, permutations = 999)
#factorfit(brayscores_ITS, samdf_ITS$int_eph, strata = samdf_ITS$substrate, permutations = 999)
```

```{r}
#myColors <- c("#9ACD32","#CD853F","#F4D166") ##56B4E9#0072B2
myColors <- c("#009E73","#E69F00","#56B4E9")
nmdsp3 <- plot_ordination(pseqtest16S, braymds_16S, type="samples", color="substrate", shape="persistence_class_state")+
  geom_point(size=2)+scale_colour_manual(values=myColors)
nmdsp3

nmds2plus16S<- plot_ordination(pseqtest16S, braymds_16S, type="samples")
nmds2plus16S$layers<-nmds2plus16S$layers[-1]
nmds2plusp16Stest<-nmds2plus16S+
  geom_point(size=3, aes(shape=persistence_class_state, color=substrate))+
  scale_colour_manual(values=myColors, name='Microhabitat:', labels=c('Epilithon', 'Leaf litter', 'Sediment'))+
  #scale_color_discrete(name='Substrate:', labels=c('Epilithic biofilms', 'Leaf litter', 'Benthic sediment'))+
  scale_shape_discrete(name='persistence_state:')+
     # scale_shape_manual(values = c(#1,
    #                                16,17,15
    #                              ))+
  #stat_ellipse(aes(x=brayscores_16S[,1],y=brayscores_16S[,2],fill=samdf_16S$substrate))+
  labs(title = "Bray-Curtis NMDS of prokaryote ASVs by substrate")+
  annotate("text", x = -0.5, y = -1.4, label = sprintf("k=2, Stress=%.4f",braymds_16S$stress), size=3.5)+
 # theme(legend.text = element_text(size=10))+
    theme(text=element_text(size=10), #change font size of all text
        axis.text=element_text(size=10), #change font size of axis text
        axis.title=element_text(size=10), #change font size of axis titles
        plot.title=element_text(size=12), #change font size of plot title
        legend.text=element_text(size=10), #change font size of legend text
        strip.text.x = element_text(size = 10),
        #legend.title=element_text(size=9),
       # legend.position = "none",
         legend.position = "right",
        plot.margin = unit(c(1, 1, 1, 2), "lines")) 
nmds2plusp16Stest


### now with legend, for publishing
nmds2plusleg16S<-nmds2plus16S+
  geom_point(size=2.5, aes(shape=persistence_class_state, fill=substrate)#, #color='black'
             )+
  scale_fill_manual(values=myColors, name='Substrate:', labels=c('Epilithon', 'Leaf litter', 'Sediment'))+
  #scale_color_discrete(name='Substrate:', labels=c('Epilithic biofilms', 'Leaf litter', 'Benthic sediment'))+
  #scale_shape_discrete(name='Site type:')+
      scale_shape_manual(name='Site state', values = c(23,24,21 #1,
                                    #16,17#,15
                                  ))+
  stat_ellipse(
    #inherit.aes = FALSE, 
               aes(x=brayscores_16S[,1],y=brayscores_16S[,2],colour=samdf_16S$substrate, linetype=(samdf_16S$persistence_class)))+
    scale_color_manual(values=myColors, name='Substrate:', labels=c('Epilithon', 'Leaf litter', 'Sediment'))+
  labs(title = "Prokaryotic community (Bray Curtis) dissimilarity")+
  annotate("text", x = 1.5, y = 1, label = sprintf("k=2, Stress=%.4f",braymds_16S$stress), size=5)+
  scale_linetype_discrete(name='Site class:')+
 # theme(legend.text = element_text(size=10))+
    theme(text=element_text(size=10), #change font size of all text
        axis.text=element_text(size=10), #change font size of axis text
        axis.title=element_text(size=10), #change font size of axis titles
        plot.title=element_text(size=14), #change font size of plot title
        legend.text=element_text(size=10), #change font size of legend text
        strip.text.x = element_text(size = 10),
        #legend.title=element_text(size=9),
        legend.position = "right",
        plot.margin = unit(c(2, 1, 2, 1), "lines"))
nmds2plusleg16S

```


#### ITS
```{r}
#generate distance matrices
bray_ITS <- phyloseq::distance(pseqtestITS, method = "bray")
#jactest <- phyloseq::distance(pseqtestITS, method="jaccard", binary=TRUE)
# NMDS
braymds_ITS <- metaMDS(bray_ITS, k=2, trymax=500, wascores = T)
brayscores_ITS <- scores(braymds_ITS)
#jacmds <- metaMDS(jactest, k=3, trymax=500, wascores = T)
#jacscores <- scores(jacmds)
braymds_ITS
#jacmds

stressplot(braymds_ITS)
#stressplot(jacmds)

factorfit(brayscores_ITS, samdf_ITS$substrate, permutations = 999)
factorfit(brayscores_ITS, samdf_ITS$WetDry, strata = samdf_ITS$substrate, permutations = 999)
factorfit(brayscores_ITS, samdf_ITS$persistence_class, strata = samdf_ITS$substrate, permutations = 999)
```

```{r}
#myColors <- c("#9ACD32","#CD853F","#F4D166")
myColors <- c("#009E73","#E69F00","#56B4E9")
nmdsp3 <- plot_ordination(pseqtestITS, braymds_ITS, type="samples", color="substrate", shape="WetDry")+
  geom_point(size=2)+scale_colour_manual(values=myColors)
nmdsp3

nmds2plusITS<- plot_ordination(pseqtestITS, braymds_ITS, type="samples")
nmds2plusITS$layers<-nmds2plusITS$layers[-1]
nmds2pluspITS<-nmds2plusITS+
  geom_point(size=2, aes(shape=persistence_class_state, color=substrate))+
  scale_colour_manual(values=myColors, name='Substrate:', labels=c('Epilithon', 'Leaf litter', 'Sediment'))+
  #scale_color_discrete(name='Substrate:', labels=c('Epilithic biofilms', 'Leaf litter', 'Benthic sediment'))+
  scale_shape_discrete(name='WetDry:')+
      scale_shape_manual(values = c(1,16,17#,15
                                  ))+
  stat_ellipse(aes(x=brayscores_ITS[,1],y=brayscores_ITS[,2],color=samdf_ITS$substrate))+
  labs(title = "NMDS of Bray Curtis dissimilarity of fungal ASV relative abundance")+
  annotate("text", x = -0.5, y = -1, label = sprintf("k=2, Stress=%.4f",braymds_ITS$stress), size=3.5)+
 # theme(legend.text = element_text(size=10))+
    theme(text=element_text(size=10), #change font size of all text
        axis.text=element_text(size=10), #change font size of axis text
        axis.title=element_text(size=10), #change font size of axis titles
        plot.title=element_text(size=12), #change font size of plot title
        legend.text=element_text(size=10), #change font size of legend text
        strip.text.x = element_text(size = 10),
        #legend.title=element_text(size=9),
        legend.position = "none",
        plot.margin = unit(c(1, 1, 1, 2), "lines")) 
nmds2pluspITS


### now with legend, for publishing
nmds2pluslegITS<-nmds2plusITS+
  geom_point(size=2.5, aes(shape=persistence_class_state, fill=substrate))+
  scale_fill_manual(values=myColors, name='Substrate:', labels=c('Epilithon', 'Leaf litter', 'Sediment'))+
  #scale_color_discrete(name='Substrate:', labels=c('Epilithic biofilms', 'Leaf litter', 'Benthic sediment'))+
  #scale_shape_discrete(name='Site type:')+
      scale_shape_manual(name='Site state', values = c(c(23,24,21
                                  )))+
  stat_ellipse(aes(x=brayscores_ITS[,1],y=brayscores_ITS[,2],color=samdf_ITS$substrate, linetype=(samdf_ITS$persistence_class)))+
  scale_colour_manual(values=myColors, name='Substrate:', labels=c('Epilithon', 'Leaf litter', 'Sediment'))+
    scale_linetype_discrete(name='Site class:')+
  labs(title = "Fungal community (Bray Curtis) dissimilarity")+
  annotate("text", x = 0.4, y = 0.45, label = sprintf("k=2, Stress=%.4f",braymds_ITS$stress), size=5)+
 # scale_linetype_discrete(name='Site class:')+
 # theme(legend.text = element_text(size=10))+
    theme(text=element_text(size=10), #change font size of all text
        axis.text=element_text(size=10), #change font size of axis text
        axis.title=element_text(size=10), #change font size of axis titles
        plot.title=element_text(size=14), #change font size of plot title
        legend.text=element_text(size=10), #change font size of legend text
        strip.text.x = element_text(size = 10),
        #legend.title=element_text(size=9),
        legend.position = "right",
        plot.margin = unit(c(2, 1, 2, 1), "lines"))
nmds2pluslegITS

```

```{r}
## load accumulation curves and venn diagrams
load("/Users/chunk/Talladega_synoptic/Analysis/Rdata/venndiagramITS.rdata")
venndiagramITS
load("/Users/chunk/Talladega_synoptic/Analysis/Rdata/ASVaccumsubITS.rdata")
ASVaccumsubITS
load("/Users/chunk/Talladega_synoptic/Analysis/Rdata/venndiagram16S.rdata")
venndiagram16S
load("/Users/chunk/Talladega_synoptic/Analysis/Rdata/ASVaccumsub16S.rdata")
ASVaccumsub16S

### now combine in a plot with
## nmds2pluslegITS
##### and 
## nmds2plusleg16S

library(cowplot)

L1<- plot_grid(ASVaccumsub16S,venndiagram16S,
          labels=c('A','B'), ncol=1, rel_heights = c(1.618,1))
R1<- plot_grid(nmds2plusleg16S,
          labels=c('C'))
L2<- plot_grid(ASVaccumsubITS,venndiagramITS,
          labels=c('D','E'), ncol=1, rel_heights = c(1.618,1))
R2<- plot_grid(nmds2pluslegITS,
          labels=c('F'))

bigp<- plot_grid(L1, R1, L2, R2, labels=NA, ncol=2, rel_widths = c(1, 1.618))
bigp

plotout <- "Beta_sixway_08.24.2025.tiff"
agg_tiff(filename=plotout, width=3200, height=3000, units="px",
         pointsize=10, res=600, compression="lzw", scaling=0.5)
bigp
invisible(dev.off())

```

# ALPHA DIVERSITY
Documentation for `estimate_richness` advises you do not rarefy data beforehand, so I am testing on the `beta` pseqtest version. 
```{r}
# 16S alpha diversity
alpha16S <- estimate_richness(pseqtest16S, split=TRUE, measures=NULL)
row.names(alpha16S)<- row.names(samdf_16S)
alpha16S <- merge(samdf_16S, alpha16S, by= 'row.names', all=TRUE)
row.names(alpha16S) <- alpha16S$Row.names 
pseq_richplot<- plot_richness(pseqtest16S, x = "substrate", color = "substrate", measures = c("Observed","Shannon"))+ geom_boxplot()
pseq_richplot

#ITS alpha diversity
alphaITS <- estimate_richness(pseqtestITS, split=TRUE, measures=NULL)
row.names(alphaITS)<- row.names(samdf_ITS)
alphaITS <- merge(samdf_ITS, alphaITS, by= 'row.names', all=TRUE)
row.names(alphaITS) <- alphaITS$Row.names 
pseq_richplot<- plot_richness(pseqtestITS, x = "substrate", color = "substrate", measures = c("Observed","Shannon"))+ geom_boxplot()
pseq_richplot
              
```

```{r}
#Set theme
theme_set(theme_bw() + theme(
              plot.title = element_text(size=16, color="black"),#,
              axis.text.x = element_text(size=12, color="black"),
              axis.text.y = element_text(size=12, color="black"),
              axis.title.x = element_text(size=12),
              axis.title.y = element_text(size=12)#,
              #legend.text = element_text(size=10),
              #legend.title = element_text(size=10),
            #  legend.position = "bottom",
            #  legend.key=element_blank(),
            #  legend.key.size = unit(0.5, "cm"),
            #  legend.spacing.x = unit(0.1, "cm"),
            #  legend.spacing.y = unit(0.1, "cm"),
              #panel.background = element_blank(), 
              #panel.border = element_rect(colour = "black", fill=NA, size=1),
              #plot.background = element_blank()))
              ))
```

```{r}
sublablist<- c('B'='Epilithic biofilms', 'L'='Leaf litter', 'S'='Benthic sediment')
sub_labeller <- function(variable,value){
  return(sublablist[value])
}
library(ggpattern) 
library(tidyverse)
library(ggpubr)
library(rstatix)
#myColors <- c("#00BFC4", "#F8766D","#7CAE00")
#names(myColors) <- levels(alpha$substrate)
#custom_colors <- scale_colour_manual(name = alpha$substrate, values = myColors)
#df <- data.frame(level = levels(alpha$substrate), outcome = c("#00BFC4", "#F8766D","#7CAE00"))

stat.test <- alpha16S %>%
wilcox_test(Observed~substrate) %>%
  adjust_pvalue(method = "bonferroni") %>%
  add_significance()
stat.test
effsize <- alpha16S %>%
  wilcox_effsize(Observed~substrate)
effsize

subalpha16S_rich <- ggplot(alpha16S, aes(x=substrate, y=Observed, fill=substrate)) +
  #facet_grid(. ~ substrate, labeller = sub_labeller)+
  #scale_x_discrete(labels=c('WET'='Wet', 'DRY'='Dry'))+
  geom_boxplot()+
  scale_x_discrete(labels = c('Epilithon','Leaf litter','Sediment'))+
    scale_fill_manual(values=myColors, name='Substrate:', labels=c('Epilithon', 'Leaf litter', 'Sediment'))+
 # geom_dotplot(binaxis = "y", stackdir = "center") +
  #geom_boxplot_pattern(aes(pattern = wet_dry), outlier.shape = NA, show.legend = FALSE) +
  #scale_pattern_manual(values= c("WET" = "none", "DRY" = "circle"))+
  #ggpattern::pattern_fill=alpha$substrate+
  #scale_pattern_fill_manual(values = c('B'="#00BFC4", 'L'="#F8766D", 'S'="#7CAE00", 'W'="#C77CFF"))+
 # scale_fill_grey(start=0.9, end=0.35) +
  ylab("ASV Richness")+
  ylim(0,6200)+
  #stat_pvalue_manual(stat.test$p.adj)+
 #geom_signif(comparisons = combn(levels(alpha$substrate),2,simplify = F),
  #            map_signif_level=TRUE)+
   geom_signif(comparisons = list(c("L","S")),y_position = 4990,
              map_signif_level=c("***"=0.001/3, "**"=0.01/3, "*"=0.05/3))+
     geom_signif(comparisons = list(c("B","S")),y_position = 5600,
            map_signif_level=c("***"=0.001/3, "**"=0.01/3, "*"=0.05/3))+
       geom_signif(comparisons = list(c("B","L")),y_position = 4900,
            map_signif_level=c("***"=0.001/3, "**"=0.01/3, "*"=0.05/3))+
 # labs(title = "ASV richness at wet versus dry sites")+
  #scale_fill_discrete(labels=c('B'='Epilithic biofilms', 'L'='Leaf litter', 'S'='Benthic sediment'))+
  #aes(fill=alpha$substrate, show.legend=FALSE)+
  theme(legend.position='none',axis.text.x = element_text(size=11), axis.title.x = element_blank(),strip.text.x = element_text(size = 11), plot.title = element_text(size=20, color="black"),
              axis.text.y = element_text(size=12, color="black"),
              axis.title.y = element_text(size=14))+
  theme(plot.margin = unit(c(2, 2, 2, 2), "lines")) 
#  geom_boxplot(aes(fill=alpha$substrate)) +
subalpha16S_rich
###############

##################################################
stat.test <- alpha16S %>%
wilcox_test(InvSimpson ~substrate) %>%
 adjust_pvalue(method = "bonferroni") %>%
  add_significance()
stat.test
effsize <- alpha16S %>%
  wilcox_effsize(InvSimpson~substrate)
effsize

subalpha16S_InvSimp <- ggplot(alpha16S, aes(x=substrate, y=InvSimpson, fill=substrate)) +
  #facet_grid(. ~ substrate, labeller = sub_labeller)+
  #scale_x_discrete(labels=c('WET'='Wet', 'DRY'='Dry'))+
  geom_boxplot()+
  scale_x_discrete(labels = c('Epilithon','Leaf litter','Sediment'))+
      scale_fill_manual(values=myColors, name='Substrate:', labels=c('Epilithon', 'Leaf litter', 'Sediment'))+
 # geom_dotplot(binaxis = "y", stackdir = "center") +
  #geom_boxplot_pattern(aes(pattern = wet_dry), outlier.shape = NA, show.legend = FALSE) +
  #scale_pattern_manual(values= c("WET" = "none", "DRY" = "circle"))+
  #ggpattern::pattern_fill=alpha$substrate+
  #scale_pattern_fill_manual(values = c('B'="#00BFC4", 'L'="#F8766D", 'S'="#7CAE00", 'W'="#C77CFF"))+
 # scale_fill_grey(start=0.9, end=0.35) +
  ylab("Inv. Simpson")+
  ylim(0,1400)+
  #stat_pvalue_manual(stat.test$p.adj)+
 #geom_signif(comparisons = combn(levels(alpha$substrate),2,simplify = F),
  #            map_signif_level=TRUE)+
   geom_signif(comparisons = list(c("L","S")),y_position = 1118,
              map_signif_level=c("***"=0.001/3, "**"=0.01/3, "*"=0.05/3))+
     geom_signif(comparisons = list(c("B","S")),y_position = 1250,
            map_signif_level=c("***"=0.001/3, "**"=0.01/3, "*"=0.05/3))+
       geom_signif(comparisons = list(c("B","L")),y_position = 900,
            map_signif_level=c("***"=0.001/3, "**"=0.01/3, "*"=0.05/3))+
  theme(legend.position='none',axis.text.x = element_text(size=11), axis.title.x = element_blank(),strip.text.x = element_text(size = 11), plot.title = element_text(size=20, color="black"),
              axis.text.y = element_text(size=12, color="black"),
              axis.title.y = element_text(size=14))+
  theme(plot.margin = unit(c(2, 2, 2, 2), "lines")) 
#  geom_boxplot(aes(fill=alpha$substrate)) +
subalpha16S_InvSimp
###############



stat.test <- alphaITS %>%
wilcox_test(Observed~substrate) %>%
  adjust_pvalue(method = "bonferroni") %>%
  add_significance()
stat.test
effsize <- alphaITS %>%
  wilcox_effsize(Observed~substrate)
effsize

subalphaITS_rich <- ggplot(alphaITS, aes(x=substrate, y=Observed, fill=substrate)) +
  #facet_grid(. ~ substrate, labeller = sub_labeller)+
  #scale_x_discrete(labels=c('WET'='Wet', 'DRY'='Dry'))+
  geom_boxplot()+
  scale_x_discrete(labels = c('Epilithon','Leaf litter','Sediment'))+
    scale_fill_manual(values=myColors, name='Substrate:', labels=c('Epilithon', 'Leaf litter', 'Sediment'))+
 # geom_dotplot(binaxis = "y", stackdir = "center") +
  #geom_boxplot_pattern(aes(pattern = wet_dry), outlier.shape = NA, show.legend = FALSE) +
  #scale_pattern_manual(values= c("WET" = "none", "DRY" = "circle"))+
  #ggpattern::pattern_fill=alpha$substrate+
  #scale_pattern_fill_manual(values = c('B'="#00BFC4", 'L'="#F8766D", 'S'="#7CAE00", 'W'="#C77CFF"))+
 # scale_fill_grey(start=0.9, end=0.35) +
  ylab("ASV Richness")+
  ylim(0,250)+
  #stat_pvalue_manual(stat.test$p.adj)+
 #geom_signif(comparisons = combn(levels(alpha$substrate),2,simplify = F),
  #            map_signif_level=TRUE)+
   geom_signif(comparisons = list(c("L","S")),y_position = 190,
              map_signif_level=c("***"=0.001/3, "**"=0.01/3, "*"=0.05/3))+
     geom_signif(comparisons = list(c("B","S")),y_position = 220,
            map_signif_level=c("***"=0.001/3, "**"=0.01/3, "*"=0.05/3))+
       geom_signif(comparisons = list(c("B","L")),y_position = 170,
            map_signif_level=c("***"=0.001/3, "**"=0.01/3, "*"=0.05/3))+
 # labs(title = "ASV richness at wet versus dry sites")+
  #scale_fill_discrete(labels=c('B'='Epilithic biofilms', 'L'='Leaf litter', 'S'='Benthic sediment'))+
  #aes(fill=alpha$substrate, show.legend=FALSE)+
  theme(legend.position='none',axis.text.x = element_text(size=11), axis.title.x = element_blank(),strip.text.x = element_text(size = 11), plot.title = element_text(size=20, color="black"),
              axis.text.y = element_text(size=12, color="black"),
              axis.title.y = element_text(size=14))+
  theme(plot.margin = unit(c(2, 2, 2, 2), "lines")) 
#  geom_boxplot(aes(fill=alpha$substrate)) +
subalphaITS_rich
###############

##################################################
stat.test <- alphaITS %>%
wilcox_test(InvSimpson ~substrate) %>%
 adjust_pvalue(method = "bonferroni") %>%
  add_significance()
stat.test
effsize <- alphaITS %>%
  wilcox_effsize(InvSimpson~substrate)
effsize

subalphaITS_InvSimp <- ggplot(alphaITS, aes(x=substrate, y=InvSimpson, fill=substrate)) +
  #facet_grid(. ~ substrate, labeller = sub_labeller)+
  #scale_x_discrete(labels=c('WET'='Wet', 'DRY'='Dry'))+
  geom_boxplot()+
  scale_x_discrete(labels = c('Epilithon','Leaf litter','Sediment'))+
      scale_fill_manual(values=myColors, name='Substrate:', labels=c('Epilithon', 'Leaf litter', 'Sediment'))+
 # geom_dotplot(binaxis = "y", stackdir = "center") +
  #geom_boxplot_pattern(aes(pattern = wet_dry), outlier.shape = NA, show.legend = FALSE) +
  #scale_pattern_manual(values= c("WET" = "none", "DRY" = "circle"))+
  #ggpattern::pattern_fill=alpha$substrate+
  #scale_pattern_fill_manual(values = c('B'="#00BFC4", 'L'="#F8766D", 'S'="#7CAE00", 'W'="#C77CFF"))+
 # scale_fill_grey(start=0.9, end=0.35) +
  ylab("Inv. Simpson")+
  ylim(0,68)+
  #stat_pvalue_manual(stat.test$p.adj)+
 #geom_signif(comparisons = combn(levels(alpha$substrate),2,simplify = F),
  #            map_signif_level=TRUE)+
   geom_signif(comparisons = list(c("L","S")),y_position = 50,
              map_signif_level=c("***"=0.001/3, "**"=0.01/3, "*"=0.05/3))+
     geom_signif(comparisons = list(c("B","S")),y_position = 59,
            map_signif_level=c("***"=0.001/3, "**"=0.01/3, "*"=0.05/3))+
       geom_signif(comparisons = list(c("B","L")),y_position = 51,
            map_signif_level=c("***"=0.001/3, "**"=0.01/3, "*"=0.05/3))+
  theme(legend.position='none',axis.text.x = element_text(size=11), axis.title.x = element_blank(),strip.text.x = element_text(size = 11), plot.title = element_text(size=20, color="black"),
              axis.text.y = element_text(size=12, color="black"),
              axis.title.y = element_text(size=14))+
  theme(plot.margin = unit(c(2, 2, 2, 2), "lines")) 
#  geom_boxplot(aes(fill=alpha$substrate)) +
subalphaITS_InvSimp
###############





######################
library(cowplot)
#plot_grid(alpha_flowstate, alpha_prcwet, alpha_DA, labels="AUTO", ncol = 1)
plotout <- "Alpha-4-panel_05-14-2025.tiff"
agg_tiff(filename=plotout, width=2200, height=2000, units="px",
         pointsize=10, res=600, compression="lzw", scaling=0.5)
plot_grid(subalpha16S_rich, subalpha16S_InvSimp, subalphaITS_rich, subalphaITS_InvSimp, labels="AUTO",label_size = 18, ncol = 2)
invisible(dev.off())

```

```{r}
## Putting all the pllots together

spacer<- ggdraw()
vnsp1<-plot_grid(spacer,venndiagram16S,spacer,rel_widths = c(0.3,1,0.3), ncol=3)
vnsp2<-plot_grid(spacer,venndiagramITS,spacer,rel_widths = c(0.3,1,0.3), ncol=3)
L1<- plot_grid(ASVaccumsub16S,vnsp1,
          labels=c('C','D'), ncol=2, rel_widths = c(1,1))
L2<- plot_grid(ASVaccumsubITS,vnsp2,
          labels=c('G','H'), ncol=2, rel_widths = c(1,1))

M1<-plot_grid(subalpha16S_rich, subalpha16S_InvSimp,
          labels=c('E','F'), ncol=2)

M2<-plot_grid(subalphaITS_rich, subalphaITS_InvSimp,
          labels=c('I','J'), ncol=2)

R1<- plot_grid(nmds2plusleg16S,
          labels=c('A'))

R2<- plot_grid(nmds2pluslegITS,
          labels=c('B'))

LM1<- plot_grid(L1, M1, rel_heights = c(0.9,1), ncol=1)
LM2<- plot_grid(L2, M2, rel_heights = c(0.9,1), ncol=1)

row1<- plot_grid(R1, LM1, ncol=2, rel_widths = c(1.2,1))
row2<- plot_grid(R2, LM2,  ncol=2, rel_widths = c(1.2,1))

#t1<- ggdraw() + draw_label("Prokaryotes (16S)",fontface = 'bold',  x = 0,  hjust = 0 ) #+theme(   plot.margin = margin(0, 0, 0, 7) )
#t2<- ggdraw() + draw_label( "Fungi (ITS)", fontface = 'bold', x = 0,  hjust = 0 ) #+ theme(   plot.margin = margin(0, 0, 0, 7) )

bigp<- plot_grid(row1, spacer, row2, labels=NA, ncol=1, rel_heights = c(1.2,0.02,1.2))
bigp

plotout <- "Overview_08.25.2025.tiff"
agg_tiff(filename=plotout, width=4500, height=3200, units="px",
         pointsize=10, res=600, compression="lzw", scaling=0.5)
bigp
invisible(dev.off())


talp<- plot_grid(R1, R2, LM1, LM2, labels=NA, ncol=2, rel_heights = c(1,1))
talp

plotout <- "Figure2_08.28.2025.tiff"
agg_tiff(filename=plotout, width=4200, height=2800, units="px",
         pointsize=10, res=600, compression="lzw", scaling=0.5)
talp
invisible(dev.off())


```

## 0.2.2
### PERMANOVA for microhabitat (substrate) types and persistence class

```{r}
### 16S by substrate
adonis2(bray_16S ~ substrate, data = samdf_16S, permutations = 99999, by="terms")
### ITS by substrate
adonis2(bray_ITS ~ substrate, data = samdf_ITS, permutations = 99999, by="terms")

## 16S intermittent versus ephemeral (strata = substrate)
adonis2(bray_16S ~ persistence_class, strata=samdf_16S$substrate, data = samdf_16S, permutations = 99999, by="terms")
## ITS intermittent versus ephemeral (strata = substrate)
adonis2(bray_ITS ~ persistence_class, strata=samdf_ITS$substrate, data = samdf_ITS, permutations = 99999, by="terms")

## wet dry at ephemeral sites only
### 16S
#pseq16S_eph<- subset_samples(pseqtest16S, int_eph=='ephemeral')
#bray_16S_eph <- phyloseq::distance(pseq16S_eph, method = "bray")
#sam_16S_eph <- data.frame(sample_data(pseq16S_eph ))
#adonis2(bray_16S_eph ~ WetDry, strata=sam_16S_eph$substrate, data = sam_16S_eph, permutations = 99999, by="terms")

### ITS
#pseqITS_eph<- subset_samples(pseqtestITS, int_eph=='ephemeral')
#bray_ITS_eph <- phyloseq::distance(pseqITS_eph, method = "bray")
#sam_ITS_eph <- data.frame(sample_data(pseqITS_eph ))
#adonis2(bray_ITS_eph ~ WetDry, strata=sam_ITS_eph$substrate, data = sam_ITS_eph, permutations = 99999, by="terms")
```


## Mantel and Partial Mantel tests for each substrate
Bray-Curtis distance of communities ~ In-stream distance between sites + Hellinger distance (for proporitions) for 11-monthh percent wet.
```{r}
set.seed(2025)
####### 16S by distance and percent wet
pseq16S_11<- subset_samples(pseqtest16S, !is.na(percentwet_11month))
## Leaf###########################################################################
### load data
pseq16S_11L<- subset_samples(pseq16S_11, substrate=="L")
sample_names(pseq16S_11L) <- pseq16S_11L@sam_data$siteId
#get sample order
sample_order <- sample_names(pseq16S_11L)
## subset in-stream distance for these samples
stic_dist <- read.csv("river_dist_stics.csv")
#
complete_pairs <- expand.grid(Row = sample_order, Column = sample_order,stringsAsFactors = FALSE)
#### merge sample list and pairwise distances
geo_complete <- merge(complete_pairs, stic_dist, by = c("Row", "Column"), all.x = TRUE)
#### set diagonal to 0
geo_complete$Distance_m[geo_complete$Row == geo_complete$Column] <- 0
missing_rev <- is.na(geo_complete$Distance_m)
geo_complete$Distance_m[missing_rev] <- geo_complete$Distance_m[match(paste(geo_complete$Column[missing_rev], geo_complete$Row[missing_rev]),paste(geo_complete$Row, geo_complete$Column))]
geo_matrix <- pivot_wider(geo_complete, names_from = Column, values_from = Distance_m, values_fill = list(Distance_m = NA)) %>% as.data.frame() %>% column_to_rownames("Row") %>% as.matrix()
geo_matrix <- geo_matrix[sample_order, sample_order]
stic_dist16SL <- as.dist(geo_matrix)
##### Now spatial matrix looks correct, on to community and prc wet
bray_16S_11L <- phyloseq::distance(pseq16S_11L, method = "bray")
sam_16S_11L <- data.frame(sample_data(pseq16S_11L))
prcw_dist <- vegdist(sam_16S_11L$percentwet_11month, method = "euclidean")

### Leaf 16S streamdist and prc_wet Mantel
mantel(bray_16S_11L, stic_dist16SL, method = "pearson", permutations = 9999)
#mantel.partial(bray_16S_11L, stic_dist16SL, prcw_dist, method = "pearson", permutations = 999)
mantel.partial(bray_16S_11L, prcw_dist, stic_dist16SL, method = "pearson", permutations = 9999)
############################################
## Epilithon
pseq16S_11B<- subset_samples(pseq16S_11, substrate=="B")
sample_names(pseq16S_11B) <- pseq16S_11B@sam_data$siteId
#get sample order
sample_order <- sample_names(pseq16S_11B)
## subset in-stream distance for these samples
stic_dist <- read.csv("river_dist_stics.csv")

complete_pairs <- expand.grid(Row = sample_order, Column = sample_order,stringsAsFactors = FALSE)
### merge sample list and pairwise distances
geo_complete <- merge(complete_pairs, stic_dist, by = c("Row", "Column"), all.x = TRUE)
### set diagonal to 0
geo_complete$Distance_m[geo_complete$Row == geo_complete$Column] <- 0
missing_rev <- is.na(geo_complete$Distance_m)
geo_complete$Distance_m[missing_rev] <- geo_complete$Distance_m[match(paste(geo_complete$Column[missing_rev], geo_complete$Row[missing_rev]),paste(geo_complete$Row, geo_complete$Column))]
geo_matrix <- pivot_wider(geo_complete, names_from = Column, values_from = Distance_m, values_fill = list(Distance_m = NA)) %>% as.data.frame() %>% column_to_rownames("Row") %>% as.matrix()
geo_matrix <- geo_matrix[sample_order, sample_order]
stic_dist16SB <- as.dist(geo_matrix)
##### Now spatial matrix looks correct, on to community and prc wet
bray_16S_11B <- phyloseq::distance(pseq16S_11B, method = "bray")
sam_16S_11B <- data.frame(sample_data(pseq16S_11B))
prcw_dist <- vegdist(sam_16S_11B$percentwet_11month, method = "euclidean")

### Epilithic biofilms 16S streamdist and prc_wet Mantel
mantel(bray_16S_11B, stic_dist16SB, method = "pearson", permutations = 9999)
#mantel.partial(bray_16S_11B, stic_dist16SB, prcw_dist, method = "pearson", permutations = 999)
mantel.partial(bray_16S_11B, prcw_dist, stic_dist16SB, method = "pearson", permutations = 9999)
################################################

## Sediment
pseq16S_11S<- subset_samples(pseq16S_11, substrate=="S")
sample_names(pseq16S_11S) <- pseq16S_11S@sam_data$siteId
#get sample order
sample_order <- sample_names(pseq16S_11S)
## subset in-stream distance for these samples
stic_dist <- read.csv("river_dist_stics.csv")

complete_pairs <- expand.grid(Row = sample_order, Column = sample_order,stringsAsFactors = FALSE)
### merge sample list and pairwise distances
geo_complete <- merge(complete_pairs, stic_dist, by = c("Row", "Column"), all.x = TRUE)
### set diagonal to 0
geo_complete$Distance_m[geo_complete$Row == geo_complete$Column] <- 0
missing_rev <- is.na(geo_complete$Distance_m)
geo_complete$Distance_m[missing_rev] <- geo_complete$Distance_m[match(paste(geo_complete$Column[missing_rev], geo_complete$Row[missing_rev]),paste(geo_complete$Row, geo_complete$Column))]
geo_matrix <- pivot_wider(geo_complete, names_from = Column, values_from = Distance_m, values_fill = list(Distance_m = NA)) %>% as.data.frame() %>% column_to_rownames("Row") %>% as.matrix()
geo_matrix <- geo_matrix[sample_order, sample_order]
stic_dist16SS <- as.dist(geo_matrix)
##### Now spatial matrix looks correct, on to community and prc wet
bray_16S_11S <- phyloseq::distance(pseq16S_11S, method = "bray")
sam_16S_11S <- data.frame(sample_data(pseq16S_11S))
prcw_dist <- vegdist(sam_16S_11S$percentwet_11month, method = "euclidean")

### Sediment 16S streamdist and prc_wet Mantel
mantel(bray_16S_11S, stic_dist16SS, method = "pearson", permutations = 9999)
#mantel.partial(bray_16S_11S, stic_dist16SS, prcw_dist, method = "pearson", permutations = 999)
mantel.partial(bray_16S_11S, prcw_dist, stic_dist16SS, method = "pearson", permutations = 9999)
################################################


####### ITS
pseqITS_11<- subset_samples(pseqtestITS, !is.na(percentwet_11month))
## Leaf###########################################################################
### load data
pseqITS_11L<- subset_samples(pseqITS_11, substrate=="L")
sample_names(pseqITS_11L) <- pseqITS_11L@sam_data$siteId
#get sample order
sample_order <- sample_names(pseqITS_11L)
## subset in-stream distance for these samples
stic_dist <- read.csv("river_dist_stics.csv")

complete_pairs <- expand.grid(Row = sample_order, Column = sample_order,stringsAsFactors = FALSE)
### merge sample list and pairwise distances
geo_complete <- merge(complete_pairs, stic_dist, by = c("Row", "Column"), all.x = TRUE)
### set diagonal to 0
geo_complete$Distance_m[geo_complete$Row == geo_complete$Column] <- 0
missing_rev <- is.na(geo_complete$Distance_m)
geo_complete$Distance_m[missing_rev] <- geo_complete$Distance_m[match(paste(geo_complete$Column[missing_rev], geo_complete$Row[missing_rev]),paste(geo_complete$Row, geo_complete$Column))]
geo_matrix <- pivot_wider(geo_complete, names_from = Column, values_from = Distance_m, values_fill = list(Distance_m = NA)) %>% as.data.frame() %>% column_to_rownames("Row") %>% as.matrix()
geo_matrix <- geo_matrix[sample_order, sample_order]
stic_distITSL <- as.dist(geo_matrix)
##### Now spatial matrix looks correct, on to community and prc wet
bray_ITS_11L <- phyloseq::distance(pseqITS_11L, method = "bray")
sam_ITS_11L <- data.frame(sample_data(pseqITS_11L))
prcw_dist <- vegdist(sam_ITS_11L$percentwet_11month, method = "euclidean")

### Leaf ITS streamdist and prc_wet Mantel
mantel(bray_ITS_11L, stic_distITSL, method = "pearson", permutations = 9999)
#mantel.partial(bray_ITS_11L, stic_distITSL, prcw_dist, method = "pearson", permutations = 999)
mantel.partial(bray_ITS_11L, prcw_dist, stic_distITSL, method = "pearson", permutations = 9999)
############################################
## Epilithon
pseqITS_11B<- subset_samples(pseqITS_11, substrate=="B")
sample_names(pseqITS_11B) <- pseqITS_11B@sam_data$siteId
#get sample order
sample_order <- sample_names(pseqITS_11B)
## subset in-stream distance for these samples
stic_dist <- read.csv("river_dist_stics.csv")

complete_pairs <- expand.grid(Row = sample_order, Column = sample_order,stringsAsFactors = FALSE)
### merge sample list and pairwise distances
geo_complete <- merge(complete_pairs, stic_dist, by = c("Row", "Column"), all.x = TRUE)
### set diagonal to 0
geo_complete$Distance_m[geo_complete$Row == geo_complete$Column] <- 0
missing_rev <- is.na(geo_complete$Distance_m)
geo_complete$Distance_m[missing_rev] <- geo_complete$Distance_m[match(paste(geo_complete$Column[missing_rev], geo_complete$Row[missing_rev]),paste(geo_complete$Row, geo_complete$Column))]
geo_matrix <- pivot_wider(geo_complete, names_from = Column, values_from = Distance_m, values_fill = list(Distance_m = NA)) %>% as.data.frame() %>% column_to_rownames("Row") %>% as.matrix()
geo_matrix <- geo_matrix[sample_order, sample_order]
stic_distITSB <- as.dist(geo_matrix)
##### Now spatial matrix looks correct, on to community and prc wet
bray_ITS_11B <- phyloseq::distance(pseqITS_11B, method = "bray")
sam_ITS_11B <- data.frame(sample_data(pseqITS_11B))
prcw_dist <- vegdist(sam_ITS_11B$percentwet_11month, method = "euclidean")

### Epilithic biofilms ITS streamdist and prc_wet Mantel
mantel(bray_ITS_11B, stic_distITSB, method = "pearson", permutations = 9999)
#mantel.partial(bray_ITS_11B, stic_distITSB, prcw_dist, method = "pearson", permutations = 999)
mantel.partial(bray_ITS_11B, prcw_dist, stic_distITSB, method = "pearson", permutations = 9999)
################################################

## Sediment
pseqITS_11S<- subset_samples(pseqITS_11, substrate=="S")
sample_names(pseqITS_11S) <- pseqITS_11S@sam_data$siteId
#get sample order
sample_order <- sample_names(pseqITS_11S)
## subset in-stream distance for these samples
stic_dist <- read.csv("river_dist_stics.csv")

complete_pairs <- expand.grid(Row = sample_order, Column = sample_order,stringsAsFactors = FALSE)
### merge sample list and pairwise distances
geo_complete <- merge(complete_pairs, stic_dist, by = c("Row", "Column"), all.x = TRUE)
### set diagonal to 0
geo_complete$Distance_m[geo_complete$Row == geo_complete$Column] <- 0
missing_rev <- is.na(geo_complete$Distance_m)
geo_complete$Distance_m[missing_rev] <- geo_complete$Distance_m[match(paste(geo_complete$Column[missing_rev], geo_complete$Row[missing_rev]),paste(geo_complete$Row, geo_complete$Column))]
geo_matrix <- pivot_wider(geo_complete, names_from = Column, values_from = Distance_m, values_fill = list(Distance_m = NA)) %>% as.data.frame() %>% column_to_rownames("Row") %>% as.matrix()
geo_matrix <- geo_matrix[sample_order, sample_order]
stic_distITSS <- as.dist(geo_matrix)
##### Now spatial matrix looks correct, on to community and prc wet
bray_ITS_11S <- phyloseq::distance(pseqITS_11S, method = "bray")
sam_ITS_11S <- data.frame(sample_data(pseqITS_11S))
prcw_dist <- vegdist(sam_ITS_11S$percentwet_11month, method = "euclidean")

### pilithic biofilms ITS streamdist and prc_wet Mantel
mantel(bray_ITS_11S, stic_distITSS, method = "pearson", permutations = 9999)
#mantel.partial(bray_ITS_11S, stic_distITSS, prcw_dist, method = "pearson", permutations = 999)
mantel.partial(bray_ITS_11S, prcw_dist, stic_distITSS, method = "pearson", permutations = 9999)
################################################

```

#### Alpha Diverity Across the Stream Continuum
Here we will use a GLM to test for effects of percent wet while accounting for a spatial covariate (mean distance of each site to all other sites, where headwater sites have higher mean distances than more central/maintstem sites). 
First, we should test alpha diversity metrics for normality:
```{r}
### checking for normality of response variables (required for gaussian)
# Shapiro-Wilk test for normality
shapiro_test <- alpha16S %>%
  group_by(substrate) %>%
  summarize(p_value = shapiro.test(Observed)$p.value)
shapiro_test

# Shapiro-Wilk test for normality
shapiro_test <- alpha16S %>%
  group_by(substrate) %>%
  summarize(p_value = shapiro.test(InvSimpson)$p.value)
shapiro_test

# Shapiro-Wilk test for normality
shapiro_test <- alphaITS %>%
  group_by(substrate) %>%
  summarize(p_value = shapiro.test(Observed)$p.value)
shapiro_test

# Shapiro-Wilk test for normality
shapiro_test <- alphaITS %>%
  group_by(substrate) %>%
  summarize(p_value = shapiro.test(InvSimpson)$p.value)
shapiro_test

```
So, in both 16S and ITS datasets, richness is normal while Inverse Simpson's Index is skewed. Therefore, we'll use Gaussian distribution for richness and Gamma for Inverse Simpson.

```{r}
#alpha16S
#alphaITS

### 16S Leaf Percent Wet & Spatial GLM
alpha16SL<- alpha16S[alpha16S$substrate=='L',]
alpha16SL<- alpha16SL[!is.na(alpha16SL$percentwet_11month),]
# 
glm_model <- glm(alpha16SL$Observed ~ alpha16SL$percentwet_11month,
                 family = gaussian)  
summary(glm_model)
glm_model <- glm(alpha16SL$InvSimpson ~ alpha16SL$percentwet_11month,
                 family = Gamma)  
summary(glm_model)
######################################################

### 16S Epilithon Percent Wet & Spatial GLM
alpha16SB<- alpha16S[alpha16S$substrate=='B',]
alpha16SB<- alpha16SB[!is.na(alpha16SB$percentwet_11month),]
# 
glm_model <- glm(alpha16SB$Observed ~ alpha16SB$percentwet_11month,
                 family = gaussian)  
summary(glm_model)
glm_model <- glm(alpha16SB$InvSimpson ~ alpha16SB$percentwet_11month,
                 family = Gamma)  
summary(glm_model)
######################################################
######################################################

### 16S Sediment Percent Wet & Spatial GLM
alpha16SS<- alpha16S[alpha16S$substrate=='S',]
alpha16SS<- alpha16SS[!is.na(alpha16SS$percentwet_11month),]
# 
glm_model <- glm(alpha16SS$Observed ~ alpha16SS$percentwet_11month,
                 family = gaussian)  
summary(glm_model)
glm_model <- glm(alpha16SS$InvSimpson ~ alpha16SS$percentwet_11month,
                 family = Gamma)  
summary(glm_model)
######################################################
######################################################

### ITS Leaf Percent Wet & Spatial GLM
alphaITSL<- alphaITS[alphaITS$substrate=='L',]
alphaITSL<- alphaITSL[!is.na(alphaITSL$percentwet_11month),]
# 
glm_model <- glm(alphaITSL$Observed ~ alphaITSL$percentwet_11month,
                 family = gaussian)  
summary(glm_model)
glm_model <- glm(alphaITSL$InvSimpson ~ alphaITSL$percentwet_11month,
                 family = Gamma)  
summary(glm_model)
######################################################

### ITS Epilithon Percent Wet & Spatial GLM
alphaITSB<- alphaITS[alphaITS$substrate=='B',]
alphaITSB<- alphaITSB[!is.na(alphaITSB$percentwet_11month),]
# 
glm_model <- glm(alphaITSB$Observed ~ alphaITSB$percentwet_11month,
                 family = gaussian)  
summary(glm_model)
glm_model <- glm(alphaITSB$InvSimpson ~ alphaITSB$percentwet_11month,
                 family = Gamma)  
summary(glm_model)
######################################################
######################################################

### ITS Sediment Percent Wet & Spatial GLM
alphaITSS<- alphaITS[alphaITS$substrate=='S',]
alphaITSS<- alphaITSS[!is.na(alphaITSS$percentwet_11month),]
# 
glm_model <- glm(alphaITSS$Observed ~ alphaITSS$percentwet_11month,
                 family = gaussian)  
summary(glm_model)
glm_model <- glm(alphaITSS$InvSimpson ~ alphaITSS$percentwet_11month,
                 family = Gamma)  
summary(glm_model)
######################################################
######################################################

```

#######################

# Leaf type effects
Now, we have presence/absence of leaf types, and can use this to make a Jaccard's distance matrix and test for effects of leaf types on microbial communities while controlling for other variables via Partial Mantel test. 

Tests for Table S2 below:
```{r}
### First select leaf samples and remake bray and spatial matrices...
set.seed(2025)
### load data
pseq16S_L<- subset_samples(pseqtest16S, substrate=="L")
sample_names(pseq16S_L) <- pseq16S_L@sam_data$siteId
#get sample order
sample_order <- sample_names(pseq16S_L)
## subset in-stream distance for these samples
stic_dist <- read.csv("river_dist_stics.csv")

complete_pairs <- expand.grid(Row = sample_order, Column = sample_order,stringsAsFactors = FALSE)
### merge sample list and pairwise distances
geo_complete <- merge(complete_pairs, stic_dist, by = c("Row", "Column"), all.x = TRUE)
### set diagonal to 0
geo_complete$Distance_m[geo_complete$Row == geo_complete$Column] <- 0
missing_rev <- is.na(geo_complete$Distance_m)
geo_complete$Distance_m[missing_rev] <- geo_complete$Distance_m[match(paste(geo_complete$Column[missing_rev], geo_complete$Row[missing_rev]),paste(geo_complete$Row, geo_complete$Column))]
geo_matrix <- pivot_wider(geo_complete, names_from = Column, values_from = Distance_m, values_fill = list(Distance_m = NA)) %>% as.data.frame() %>% column_to_rownames("Row") %>% as.matrix()
geo_matrix <- geo_matrix[sample_order, sample_order]
stic_dist16SL <- as.dist(geo_matrix)
##### Now spatial matrix looks correct, on to community and prc wet
bray_16S_L <- phyloseq::distance(pseq16S_L, method = "bray")
sam_16S_L <- data.frame(sample_data(pseq16S_L))
#prcw_dist <- vegdist(sam_16S_L$percentwet_month, method = "euclidean")

### Now, extract leaf presence absence, and remove singleton leaf types
leaf_pres.abs<- sam_16S_L[,22:41]
leaf_pres.abs<- leaf_pres.abs[,colSums(leaf_pres.abs)>1]
leaf_pres.abs<- leaf_pres.abs[,colnames(leaf_pres.abs)!='Unknown']

## making Jaccard's distance matrix from leaf type presence/absence in each composite leaf sample
leafdist<- vegdist(leaf_pres.abs, method = "jaccard")

### Leaf 16S streamdist and prc_wet Mantel

# TABLE S2A
mantel(leafdist, stic_dist16SL, method = "pearson", permutations = 9999)

# TABLE S2C
mantel(bray_16S_L, leafdist, method = "pearson", permutations = 9999)

# TABLE S2D
#mantel.partial(bray_16S_L, stic_dist16SL, prcw_dist, method = "pearson", permutations = 999)
mantel.partial(bray_16S_L, leafdist, stic_dist16SL, method = "pearson", permutations = 99999)

```

```{r}
### First select leaf samples and remake bray and spatial matrices...
set.seed(2025)
### load data
pseqITS_L<- subset_samples(pseqtestITS, substrate=="L")
sample_names(pseqITS_L) <- pseqITS_L@sam_data$siteId
#get sample order
sample_order <- sample_names(pseqITS_L)
## subset in-stream distance for these samples
stic_dist <- read.csv("river_dist_stics.csv")

complete_pairs <- expand.grid(Row = sample_order, Column = sample_order,stringsAsFactors = FALSE)
### merge sample list and pairwise distances
geo_complete <- merge(complete_pairs, stic_dist, by = c("Row", "Column"), all.x = TRUE)
### set diagonal to 0
geo_complete$Distance_m[geo_complete$Row == geo_complete$Column] <- 0
missing_rev <- is.na(geo_complete$Distance_m)
geo_complete$Distance_m[missing_rev] <- geo_complete$Distance_m[match(paste(geo_complete$Column[missing_rev], geo_complete$Row[missing_rev]),paste(geo_complete$Row, geo_complete$Column))]
geo_matrix <- pivot_wider(geo_complete, names_from = Column, values_from = Distance_m, values_fill = list(Distance_m = NA)) %>% as.data.frame() %>% column_to_rownames("Row") %>% as.matrix()
geo_matrix <- geo_matrix[sample_order, sample_order]
stic_distITSL <- as.dist(geo_matrix)
##### Now spatial matrix looks correct, on to community and prc wet
bray_ITS_L <- phyloseq::distance(pseqITS_L, method = "bray")
sam_ITS_L <- data.frame(sample_data(pseqITS_L))
#prcw_dist <- vegdist(sam_ITS_L$percentwet_month, method = "euclidean")

### Now, extract leaf presence absence, and remove singleton leaf types
leaf_pres.abs<- sam_ITS_L[,22:41]
leaf_pres.abs<- leaf_pres.abs[,colSums(leaf_pres.abs)>1]
leaf_pres.abs<- leaf_pres.abs[,colnames(leaf_pres.abs)!='Unknown']

## making Jaccard's distance matrix from leaf type presence/absence in each composite leaf sample
leafdist<- vegdist(leaf_pres.abs, method = "jaccard")


# TABLE S2F
mantel(bray_ITS_L, leafdist, method = "pearson", permutations = 9999)

# TABLE S2G
mantel.partial(bray_ITS_L, leafdist, stic_distITSL, method = "pearson", permutations = 99999)

```

# checking water permanence versus leaf types
```{r}
### First select leaf samples and remake bray and spatial matrices...
set.seed(2025)
pseq16S_11<- subset_samples(pseqtest16S, !is.na(percentwet_11month))
### load data
pseq16S_L<- subset_samples(pseq16S_11, substrate=="L")
sample_names(pseq16S_L) <- pseq16S_L@sam_data$siteId
#get sample order
sample_order <- sample_names(pseq16S_L)
## subset in-stream distance for these samples
stic_dist <- read.csv("river_dist_stics.csv")

complete_pairs <- expand.grid(Row = sample_order, Column = sample_order,stringsAsFactors = FALSE)
### merge sample list and pairwise distances
geo_complete <- merge(complete_pairs, stic_dist, by = c("Row", "Column"), all.x = TRUE)
### set diagonal to 0
geo_complete$Distance_m[geo_complete$Row == geo_complete$Column] <- 0
missing_rev <- is.na(geo_complete$Distance_m)
geo_complete$Distance_m[missing_rev] <- geo_complete$Distance_m[match(paste(geo_complete$Column[missing_rev], geo_complete$Row[missing_rev]),paste(geo_complete$Row, geo_complete$Column))]
geo_matrix <- pivot_wider(geo_complete, names_from = Column, values_from = Distance_m, values_fill = list(Distance_m = NA)) %>% as.data.frame() %>% column_to_rownames("Row") %>% as.matrix()
geo_matrix <- geo_matrix[sample_order, sample_order]
stic_dist16SL <- as.dist(geo_matrix)
##### Now spatial matrix looks correct, on to community and prc wet
bray_16S_L <- phyloseq::distance(pseq16S_L, method = "bray")
sam_16S_L <- data.frame(sample_data(pseq16S_L))
prcw_dist <- vegdist(sam_16S_L$percentwet_11month, method = "euclidean")

### Now, extract leaf presence absence, and remove singleton leaf types
leaf_pres.abs<- sam_16S_L[,22:41]
leaf_pres.abs<- leaf_pres.abs[,colSums(leaf_pres.abs)>1]
leaf_pres.abs<- leaf_pres.abs[,colnames(leaf_pres.abs)!='Unknown']

## making Jaccard's distance matrix from leaf type presence/absence in each composite leaf sample
leafdist<- vegdist(leaf_pres.abs, method = "jaccard")

## CHECKING leaf type versus percent wet

### Leaf 16S leaf types and prc_wet Mantel
# TABLE S2B
mantel(leafdist, prcw_dist, method = "pearson", permutations = 9999)

mantel(bray_16S_L, leafdist, method = "pearson", permutations = 9999)

#mantel.partial(bray_16S_L, stic_dist16SL, prcw_dist, method = "pearson", permutations = 999)

# TABLE S2E
mantel.partial(bray_16S_L, leafdist, prcw_dist, method = "pearson", permutations = 99999)

```

```{r}
### First select leaf samples and remake bray and spatial matrices...
set.seed(2025)
### load data
pseqITS_11<- subset_samples(pseqtestITS, !is.na(percentwet_11month))
pseqITS_L<- subset_samples(pseqITS_11, substrate=="L")
sample_names(pseqITS_L) <- pseqITS_L@sam_data$siteId
#get sample order
sample_order <- sample_names(pseqITS_L)
## subset in-stream distance for these samples
stic_dist <- read.csv("river_dist_stics.csv")

complete_pairs <- expand.grid(Row = sample_order, Column = sample_order,stringsAsFactors = FALSE)
### merge sample list and pairwise distances
geo_complete <- merge(complete_pairs, stic_dist, by = c("Row", "Column"), all.x = TRUE)
### set diagonal to 0
geo_complete$Distance_m[geo_complete$Row == geo_complete$Column] <- 0
missing_rev <- is.na(geo_complete$Distance_m)
geo_complete$Distance_m[missing_rev] <- geo_complete$Distance_m[match(paste(geo_complete$Column[missing_rev], geo_complete$Row[missing_rev]),paste(geo_complete$Row, geo_complete$Column))]
geo_matrix <- pivot_wider(geo_complete, names_from = Column, values_from = Distance_m, values_fill = list(Distance_m = NA)) %>% as.data.frame() %>% column_to_rownames("Row") %>% as.matrix()
geo_matrix <- geo_matrix[sample_order, sample_order]
stic_distITSL <- as.dist(geo_matrix)
##### Now spatial matrix looks correct, on to community and prc wet
bray_ITS_L <- phyloseq::distance(pseqITS_L, method = "bray")
sam_ITS_L <- data.frame(sample_data(pseqITS_L))
#prcw_dist <- vegdist(sam_ITS_L$percentwet_month, method = "euclidean")

### Now, extract leaf presence absence, and remove singleton leaf types
leaf_pres.abs<- sam_ITS_L[,22:41]
leaf_pres.abs<- leaf_pres.abs[,colSums(leaf_pres.abs)>1]
leaf_pres.abs<- leaf_pres.abs[,colnames(leaf_pres.abs)!='Unknown']

## making Jaccard's distance matrix from leaf type presence/absence in each composite leaf sample
leafdist<- vegdist(leaf_pres.abs, method = "jaccard")

# TABLE S2H
mantel.partial(bray_ITS_L, leafdist, prcw_dist, method = "pearson", permutations = 99999)
```

